## Outline for today

- The standard normal distribution

- Transformations

    - Centering

    - Scaling

    - The *z*-transform

- Making comparisons

  - Comparing groups

  - Comparing across groups

  - Making comparisons with the sampling distribution 

---

## The shape of things


```{r}
#| echo: false
#| message: false
require(ggplot2)
require(cowplot)
set.seed(124)
df <- tibble::tibble(x = rnorm(n = 1000, mean = 0, sd = 1) |>
scale() |> as.numeric())
df_og <- df
df$x <- df$x * 10
df$x <- df$x + 165
df_m <- df$x |> mean()
df_s <- df$x |> sd()
range <- glue::glue("{round(df_m - df_s  * 1)}--{round(df_m + df_s * 1)}")
p <- ggplot(df, aes(x = x)) +
  geom_histogram(binwidth = 5, boundary = 130) +
  labs(x = "height in cm") +
  theme_cowplot() +
  NULL
```

- If we measured the height of `r df$x |> length()` women and plotted the
  values then we might get something like @fig-height-histogram. 

- Most heights are in the `r range` centimetre range. 

- The distribution is roughly symmetrical around its mean (`r round(df_m)` cm)
  and it has a shape characteristic of a normal distribution.

```{r}
#| echo: false
#| label: fig-height-histogram
#| fig.cap: |
#|    Distribution of heights in a sample of 1000 women. Not real
#|    data.
p
```

---

## The shape of things

::::{.columns}

:::{.column width="50%"}

- Of course the plot in @fig-height-histogram doesn't look exactly like a normal
distribution 

- But if we measured more and more people (e.g., 100, 000 people) then we might
  get something like @fig-height-histogram2

- @fig-height-histogram2 also shows the corresponding normal distribution with
  a mean of `r df_m` and a standard deviation of `r df_s`

- Although the **normal distribution** is an idealisation, or an abstraction,
  we can use it to do some very useful things
:::

:::{.column width="50%"}

```{r}
#| echo: false
#| label: fig-height-histogram2
#| fig.cap: |
#|    Distribution of heights in a sample of 100,000 women (Not real
#|    data) and the corresponding normal distribution
df <- tibble::tibble(x = rnorm(n = 100000, mean = 0, sd = 1) |>
scale() |> as.numeric())
df$x <- df$x * 10
df$x <- df$x + 165
df_m <- df$x |> mean()
df_s <- df$x |> sd()
ideal <- \(x) (dnorm(x, df_m, df_s) / dnorm(df_m, df_m, df_s)) * ((195 / 5) * 100)
ggplot(df, aes(x)) +
  geom_histogram(binwidth = 1, boundary = 130) +
  geom_density(col = "grey20", size = 2, aes(y = ..count.. )) +
  geom_function(fun = ideal, col = "blue", size = 2) +
  labs(y = "count", x = "height in cm") +
  theme_cowplot()
```

:::

::::

---

## The standard normal distribution 


- In lecture 8, I said that **two parameters**, $\mu$ and $\sigma$ changed
  where the normal distribution was centred and how spread out it was

- When $\mu = 0$ and $\sigma = 1$, then this distribution is called the
  **standard normal distribution**

- I said that changing these values didn't change the **relative position**
  of points on the plot. The overall shape remains the same

- All normal distributions have the same **overall shape** as the **standard
  normal distribution** even if they're centered in a different place and are
  more or less spread out

To see what I mean by this, we'll take out heights of 1000 people, but 
instead of displaying them in centimetres we'll display them in metres 

---


```{r}
#| echo: false
#| label: fig-height-histogram-new
#| fig-cap: |
#|    Distribution of heights in a sample of 1000 women. Not real
#|    data.
#| fig-subcap:
#|    - "Measured in centimetres"
#|    - "Measured in metres"
#| layout-ncol: 2
# require(ggplot2)
# require(plotshow)
require(cowplot)
set.seed(124)
df <- tibble::tibble(x = rnorm(n = 1000, mean = 0, sd = 1) |>
scale() |> as.numeric())
df$x <- df$x * 10
df$x <- df$x + 165
df_m <- df$x |> mean()
df_s1 <- df$x |> sd()
range <- glue::glue("{round(df_m - df_s  * 1)}--{round(df_m + df_s * 1)}")
ggplot(df, aes(x = x)) +
  geom_histogram(binwidth = 5, boundary = 130) +
  labs(x = "height in cm") +
  theme_cowplot() +
  NULL


set.seed(124)
df <- tibble::tibble(x = rnorm(n = 1000, mean = 0, sd = 1) |>
scale() |> as.numeric())
df$x <- df$x * (10 / 100)
df$x <- df$x + (165 / 100)
df_m <- df$x |> mean()
df_s2 <- df$x |> sd()
range <- glue::glue("{round(df_m - df_s  * 1)}--{round(df_m + df_s * 1)}")
ggplot(df, aes(x = x)) +
  geom_histogram(binwidth = 5 / 100, boundary = 130 / 100) +
  labs(x = "height in m") +
  theme_cowplot() +
  NULL
```

Changing the scale on which you're measured doesn't actually change
your height relative to other people

---

- The distribution in @fig-height-histogram-new-1 has a standard deviation
of `r df_s1` 

- The distribution in @fig-height-histogram-new-2 has a standard deviation of
  `r df_s2`.

But as you can see, they're the same distributions---they're just displayed on
**different scales** (centimetres versus metres).


\
Changing the **scale** changes the **standard deviation**. This is why
the **standard deviation** is sometimes referred to as the **scale
parameter** for the distribution.

---


-  Apart from changing the scale, we can also change where the distribution
is centred


```{r}
#| echo: false
#| label: fig-height-histogram-new-location
#| fig-cap: |
#|    Distribution of heights in a sample of 1000 women. Not real
#|    data.
#| fig-subcap:
#|    - "Measured in centimetres"
#|    - "Measured in difference from the average height"
#| layout-ncol: 2
# require(ggplot2)
# require(plotshow)
require(cowplot)
set.seed(124)
df <- tibble::tibble(x = rnorm(n = 1000, mean = 0, sd = 1) |>
scale() |> as.numeric())
df$x <- df$x * 10
df$x <- df$x + 165 
df_m <- df$x |> mean()
df_s1 <- df$x |> sd()
range <- glue::glue("{round(df_m - df_s  * 1)}--{round(df_m + df_s * 1)}")
ggplot(df, aes(x = x)) +
  geom_histogram(binwidth = 5, boundary = 130) +
  labs(x = "height in cm") +
  theme_cowplot() +
  NULL


set.seed(124)
df <- tibble::tibble(x = rnorm(n = 1000, mean = 0, sd = 1) |>
scale() |> as.numeric())
df$x <- df$x * (10 )
df$x <- df$x + (165 - 165)
df_m <- df$x |> mean()
df_s2 <- df$x |> sd()
range <- glue::glue("{round(df_m - df_s  * 1)}--{round(df_m + df_s * 1)}")
ggplot(df, aes(x = x)) +
  geom_histogram(binwidth = 5,  boundary = 130 - 165) +
  labs(x = "difference from average height in cm") +
  theme_cowplot() +
  NULL
```

- In @fig-height-histogram-new-location-1 we can see the same distribution
as before. In @fig-height-histogram-new-location-2 we can see a
distribution that is now centred at 0. 

---

### The normal distribution

::::{.columns}

:::{.column width="50%"}

- The fact that the relative positions of points don't change is a useful
property. 

- For the standard normal distribution, ~68% of the distribution
falls between -1 and +1. 

- Put into relative terms this is ±1 $\sigma$ from $\mu$.


```{ojs}
texmd`- The highlighted region covers the region from $\mu$ -${s} $\sigma$ to $\mu$ + ${s} $\sigma$

- Or ${mean_value - s * sd_value} to ${mean_value + s * sd_value} 

- This covers ~${coverage(s)}% of the distribution not what the value of $\mu$ and $\sigma$

`
```

```{ojs}
function coverage(s) {
let cov = jstat.normal.cdf(s, 0, 1) * 2 - 1
return Math.round(cov * 100)
}
```


:::

:::{.column width="50%"}


```{ojs}
normal_plot_output = Plot.plot({
  x: {
    grid: true,
    domain: [-4, 4]
  },
  y: {
    grid: true,
    domain: [0, 0.8]
  },
  marks: [
   Plot.line(
     normal_plot(
       mean_value - sd_value * s > -4 ? mean_value - sd_value * s : -4,
       mean_value + sd_value * s < 4 ? mean_value + sd_value * s : 4,
       mean_value,
       sd_value
     ),
     {
       x: "x",
       y: "y",
       strokeWidth: 1,
       fill: "blue",
       opacity: true ? 0.5 : 0
     }
   ),
   Plot.line(fill_limits(s), {
     x: "x",
     y: "y",
     fill: "blue",
     strokeWidth: 1,
     opacity: true ? 0.5 : 0
   }),
    Plot.line(normal_plot(-4, 4, mean_value, sd_value), {
      x: "x",
      y: "y",
      strokeWidth: 4
    }),
    Plot.ruleY([0])
  ]
})
```

```{ojs}
//| panel: input
normal_sliders = htl.html`${mean_value_slider}${sd_value_slider}`
viewof s = Inputs.range([0, 3], { step: 0.5, value: 1 })
```


:::

::::


---

## Transformations

- In @fig-height-histogram-new and @fig-height-histogram-new-location we saw 
that we could **transform** a variable so that it had a new **location** (mean)
or **scale** (standard deviation) without changing the shape

- These two kinds of **transformations** are known as **centering** and **scaling*

### Centering 


- To centre a set of measurements, you **subtract** a fixed value from each observation in
the dataset. 

- This has the effect of shifting the distribution of the variable _along the
  x-axis_

- You can technically centre a variable by subtracting _any value_ from it but
  the most frequently used method is **mean-centring**

This is shown in @eq-center, below:

$$x_i - \bar{x}$${#eq-center}

---

### Mean centering

- Mean centering a variable shifts it so that the new mean is at the zero point

- The individual values of a mean-centred variable tell us how far that
  observation is from the mean of the entire set of measures

- It doesn't alter the **shape** of the distribution, or change the scale that
  it's measured on

- It only changes the interpretation of the values to, for example,
  _differences from the mean_

---

### Scaling

- Scaling is performed by **dividing** each observation by some fixed value

-  This has the effect of stretching or compressing the variable _along the
   x-axis_

- You can scale a variable by dividing it by **any** value

    - We created @fig-height-histogram-new-2 by taking the values in
      @fig-height-histogram-new-1 and dividing them by 100 (to transform cm to
      m)

- But typically scaling is done by dividing values by the standard deviation of
  the dataset. 

This is shown in @eq-scale, below:

$$\frac{x_i}{s}$${#eq-scale}

- Scaling doesn't change the fundamental shape of the variable's distribution

- But after scaling the data by the standard deviation the values would now be
  measured in _units of sd_

---

### The *z* transform

- The combination of first mean-centering a variable and then scaling it by its
  standard deviation is known as the **z**-transform


The formula for this is shown in @eq-z, below:

$$z(x) = \frac{x_i - \bar{x}}{s}$${#eq-z}


---

### The *z* transform

```{r}
set.seed(555)
d <- rnorm(10, 5, 2) |> round()
```

::::{.columns}

:::{.column width="50%"}

- The 10 values in @tbl-z have a mean of `r mean(d)` and a standard deviation
of `r sd(d) |> round(2)`. 

- To **z**transform the data in @tbl-z. We would do the following steps:

 1.  We'd subtract `r mean(d)` from each value and put them in the **Centred** column 

  2. Then we'd divide each value in **Centred** by `r sd(d) |> round(2)`

- We can now interpret the data in terms of **distance from the mean in units
  of standard deviation**

The **z** transform will come in handy when it comes to making comparisons

:::

:::{.column width="50%"}


```{r}
#| label: tbl-z
#| tbl-cap: z transformed data
tibble::tibble(r = d) |>
  dplyr::mutate(centred = d - mean(d)) |>
  dplyr::mutate(scaled = centred / sd(d)) |>
  magrittr::set_colnames(
c("Raw values",
"Centred",
"Scaled")) |>
knitr::kable(digits = 2)
```

:::

::::


---

## Comparing groups

```{r}
sport_data <- tibble::tribble(
  ~group, ~mean, ~sd,
  "amateur", 500, 30,
  "sport", 460, 15
)


ama_mean <- sport_data |>
  dplyr::filter(group == "amateur") |>
  dplyr::pull(mean)
spo_mean <- sport_data |>
  dplyr::filter(group == "sport") |>
  dplyr::pull(mean)

N <- 500

set.seed(123)
sport_tib <- purrr::pmap_df(sport_data, function(group, mean, sd) {
  d <- rnorm(N, mean, sd) |> scale() |> as.numeric()
  tibble::tibble(group = group, x = mean + (d * sd))
})
```

::::{.columns}

:::{.column width="50%"}

In the context of **quantitative research** we're often looking at the
**average difference** in a variable between groups

In the @fig-sport-histogram we can see measurements from a reaction time task. 

- Amateurs sportspeople have a mean reaction time of `r ama_mean` ms and
  professionals have a mean reaction time of `r spo_mean` ms.

- There is overlap between the two groups, but there is a difference between
  the **averages**

- To quantify the difference, just **subtract the mean of
  one group from the mean of the other**^[In later years you'll learn how to
  quantify differences like this in standardised units---that is, units of
  standard deviation]

:::

:::{.column width="50%"}

```{r}
#| echo: false
#| warning: false
#| label: fig-sport-histogram
#| fig.align: center
#| fig.cap: |
#|    Distribution of reaction times in a sample of amateur (green) and
#|    500 professional (blue) sportspeople. Group means are indicated
#|    with the vertical lines.
require(ggplot2)

ggplot2::ggplot(sport_tib, aes(x = x, 
  fill = group,
  group = group,
  alpha = group,
  color = group,
  size = group)) +
  geom_histogram(bins = 50, boundary = 100, position = "identity") +
  scale_alpha_manual(values = c(1, 0.7), guide = "none") +
  scale_fill_manual(values = c("seagreen", "blue"), guide = "none") +
  scale_color_manual(values = c("white", "white"), guide = "none") +
  scale_size_manual(values = c(.2, .2), guide = "none") +
  geom_vline(xintercept = ama_mean, color = "seagreen", size = 1.5) +
  geom_vline(xintercept = spo_mean, color = "blue", size = 1.5) +
  ggplot2::annotate(geom = "text", x = ama_mean, y = 60,
  label = expression(bar('X')[amateur]), hjust = -0.3, color = "seagreen") +
  ggplot2::annotate(geom = "text", x = spo_mean, y = 60,
  label = expression(bar('X')[pro]), hjust = -0.3, color = "blue") +
  theme_cowplot() +
  labs(x = "reaction time (ms)", y = "count") +
  NULL

```


- The mean **difference** is just `r ama_mean`ms - `r spo_mean`ms = `r ama_mean - spo_mean`ms.


:::

::::


---

## Comparing across groups

- In the previous example the comparisons were easy because the measurements were
  on the same scale (milliseconds)

- But let's say that you want to compare two children on a puzzle completion task

    - One child is 8 years old, and the other is 14 years old

    - They do slightly different versions of the task and the tasks are score
      differently 

- Because we have two different tests that might have a different number of
  items etc we can't just compare the **raw numbers** to see which is bigger


Let's look at an example...

---

### Comparing across groups


```{r}
#| echo: false
stats <- tibble::tribble(
  ~age, ~mean, ~sd,
  "8", 80, 2,
  "14", 120, 8,
  )

a_z <- 3
b_z <- .5
a_child <- stats$mean[1] + (stats$sd[1] * a_z)
b_child <- stats$mean[2] + (stats$sd[2] * b_z)


```

Let's take two children:

- Ahorangi is 8 years old and scored `r a_child` on the task

- Benjamin is 14 years old and scored `r b_child` on the task

We can easily tell that Benjamin's score is higher than Ahorangi's score

But the scores are not directly comparable...so what do we do?

- We have to look at how each performed **relative** to their age groups.

- Is Ahorangi better performing relative to 8-year-olds than Benjamin is
  relative to 14-year-olds?

- To answer this question we can use the **z**-transform

---

### Comparing across groups

To do the **z**-transform we need to know the **mean** and **standard deviation** 
for each age group

::::{.columns}

:::{.column width="50%"}

```{r}
#| echo: false
#| label: tbl-age-means
#| tbl-cap: |
#|      Means and Standard deviations for the 8-year-old and
#|      14-year-old age groups
stats |>
  dplyr::mutate(age = dplyr::case_when(
    age == 8 ~ "8-year-olds",
    age == "14" ~ "14-year-olds")) |>
  knitr::kable(col.names = c("Age group", "Mean", "Standard deviation"))
```

:::

:::{.column width="50%"}

We can calculate Ahorangi's z-score as follows:

$$`r a_z` = \frac{`r a_child` - `r stats$mean[1]`}{`r stats$sd[1]`}$$

And for Benjamin:


$$`r b_z` = \frac{`r b_child` - `r stats$mean[2]`}{`r stats$sd[2]`}$$

:::

::::

That means, that Ahorangi, despite having a lower score, actually scored very
high for an 8-year-old. Benjamin, on the other hand, only scored a little
higher than the average 14-year-old.

---

## Making comparisons with the sampling distribution

The final comparison we'll talk about is comparisons against the **sampling distribution**

- From last week we learned that the **sampling distribution of the mean** will be

    - Centred at the **population mean**

    - Have a standard deviation equal to the **standard error of the mean**

- But remember, we don't know the value of the **population mean**, so we won't
  actually know what the **sampling distribution** looks like

Although we don't know the value of the **population mean** we can generate a
**hypothesis** about what we think the population mean might be...

We can then generate a **hypothetical sampling distribution** based on our
**hypothesised value** of the population mean

---

### Making comparisons with the sampling distribution

```{r}
#| echo: false
#| message: false
require(tidyverse)
# require(plotshow)
set.seed(123)
n <- 50
face_celeb <- rnorm(n, 500, 50)
face_family <- rnorm(n, 520, 45)
subject_id <- seq(1, n)
df <- tibble::tibble(face_celeb, face_family, subject_id)

df |>
  tidyr::pivot_longer(1:2,
    names_to = "condition",
          values_to = "rt")|>
  dplyr::group_by(subject_id) |>
  dplyr::arrange(condition) |>
  dplyr::summarise(diff = diff(rt)) |>
  dplyr::ungroup() -> df

sum_stats <- df  |>
  dplyr::summarise(
    mean_rt = mean(diff),
    sd_rt = sd(diff),
    n = n(),
    se = sd_rt / sqrt(n)
)




```

I'll make it concrete with an example:

- Let's say I get a group of people to perform a task where they have to try
  and quickly recognise two sets of faces. Either famous faces or faces of
  their family members.

- I find that the mean difference between these two conditions is `r sum_stats$mean_rt |> round(2)`ms

- But this is just the difference in my **sample**. The population mean
  difference might be some other value

- Although we don't know the population mean, we could hypothesise that it is
  100 ms, 50 ms, 0 ms, or some other value. Let's just pick 0 ms for now. 

- Now we can generate a sampling distribution using our hypothesised
  **population mean** and the **standard error of the mean** we estimate from
  the sample (let's say it's `r round(sum_stats$sd_rt / sqrt(sum_stats$n), 2)`)

---

### Making comparisons with the sampling distribution

In @fig-sampling we can see what the sampling distribution would
look like if the population mean were 0.

::::{.columns}

:::{.column width="50%"}

- We can compare our particular sample mean of 
  `r sum_stats$mean_rt |> round(2)`ms to the sampling distribution

- Because the sampling distribution is a normal distribution we know that ~68% of
  the time the sample means will fall between ±1 SEM of the population mean 
  (`r round(sum_stats$se, 2) * -1`ms to `r round(sum_stats$se, 2)`ms)

- And ~95% of the time sample means will fall between `r round(sum_stats$se, 2) * -2`ms
and `r round(sum_stats$se, 2) * 2`ms. 



:::

:::{.column width="50%"}

```{r}
#| label: fig-sampling
#| fig-cap: !expr glue::glue("The sampling distribution with a mean of 0 and a SEM of {sum_stats$se |> round(2)}")
require(tidyverse)

args <- sum_stats |>
  magrittr::set_colnames(
    c("mean", "sd", "n","se")
) |>
  dplyr::mutate(sd = sd / sqrt(n)) |>
  dplyr::mutate(mean = 0) |>
  dplyr::select(-n, -se) |> as.list()



ggplot() +
  geom_function(fun = dnorm, args = args) +
  xlim(args$mean - 4 * args$sd, args$mean + 4 * args$sd) +
  cowplot::theme_cowplot() +
  labs(y = NULL, x = "sample mean")

```

:::

::::


For our particular mean we see that it falls `r round(sum_stats$mean_rt / sum_stats$se, 2)` SEM 
from our hypothesised population mean

What can we make of this?

---

### Making comparisons with the sampling distribution

- We can conclude that if the population mean were in fact 0 then we have observed something rare

- If the population mean were in fact 0, then it would be rare for a sample mean to be that 
far away from the population mean

Observing something rare doesn't tell us that our hypothesis is wrong 

Rare things happen all the time!

- But if we were to run our experiments again and again, and we continued to observe
rare events then we would probably have a good reason to update our hypothesis. 

This process of comparing our sample to the sampling distribution is known as
*null hypothesis significance testing*

And it will be a major topic that you'll cover next year. 






























```{ojs}
vega = require("https://cdn.jsdelivr.net/npm/vega@5/build/vega.js")
```

```{ojs}
jstat = require("https://bundle.run/jstat@1.9.4")
```

```{ojs}
import { dist } from "@ljcolling/wasm-distributions"
```

```{ojs}
sd_value_slider = Inputs.range([0.5, 2], {
  value: 1,
  step: 0.25,
  label: htl.html`standard deviation <br />&#x3C3;`
})
```

```{ojs}
import { texmd } from "@kelleyvanevert/katex-within-markdown"
```

```{ojs}
mean_value_slider = Inputs.range([-3, 3], {
  value: 0,
  step: 0.25,
  label: htl.html`mean<br />  &#x3BC`
})
```

```{ojs}
sd_value = Generators.input(sd_value_slider)
```

```{ojs}
mean_value = Generators.input(mean_value_slider)
```

```{ojs}
// jStat.normal.pdf( x, mean, std )
normal_plot = (min, max, mean, sd) => {
  // jStat.normal.pdf(x, mean, sd)

  return d3.ticks(min, max, 501).map((v) => {
    return {
      x: v,
      y: dnorm(v, mean, sd)
    };
  });
}
```

```{ojs}
skew_normal_plot = (min, max, alpha) => {
  // jStat.normal.pdf(x, mean, sd)

  return d3.ticks(min, max, 201).map((v) => {
    return {
      x: v,
      y: dsn(v, alpha)
    };
  });
}
```

```{ojs}
dsn = (x, alpha) => {
  // set the defaults

  const xi = 0;
  const omega = 1;
  const tau = 0;

  let z = (x - xi) / omega;

  let logN = -Math.log(Math.sqrt(2 * Math.PI)) - 0 - Math.pow(z, 2) / 2;

  let logS = Math.log(
    jStat.normal.cdf(tau * Math.sqrt(1 + Math.pow(alpha, 2)) + alpha * z, 0, 1)
  );

  let logPDF = logN + logS - Math.log(jStat.normal.cdf(tau, 0, 1));

  return Math.exp(logPDF);
}
```

```{ojs}
// import jStat library
jStat = require("jStat")
```

```{ojs}
kurtosis = {
  return {
    uniform: -(6 / 5),
    raised_cosine: (6 * (90 - Math.PI ** 4)) / (5 * (Math.PI ** 2 - 6) ** 2),
    standard_normal: 0,
    t_dist30: 6 / (30 - 4),
    t_dist20: 6 / (20 - 4),
    t_dist10: 6 / (10 - 4),
    t_dist7: 6 / (7 - 5),
    t_dist5: 6 / (5 - 4)
  };
}
```

```{ojs}
kurtosis_values = Object.values(kurtosis).map((v) => Math.round(v * 100) / 100)
```

```{ojs}
dists = {
  return {
    raised_cosine: d3.ticks(-3, 3, 500).map((v) => {
      return {
        x: v,
        y: dist.raised_cosine(v, 0, 2.5)
      };
    }),
    standard_normal: d3.ticks(-3, 3, 500).map((v) => {
      return {
        x: v,
        y: dnorm(v, 0, 1)
      };
    }),
    t_dist30: d3.ticks(-3, 3, 500).map((v) => {
      return {
        x: v,
        y: dist.dt(v, 30)
      };
    }),
    t_dist20: d3.ticks(-3, 3, 500).map((v) => {
      return {
        x: v,
        y: dist.dt(v, 20)
      };
    }),
    t_dist10: d3.ticks(-3, 3, 500).map((v) => {
      return {
        x: v,
        y: dist.dt(v, 10)
      };
    }),
    t_dist7: d3.ticks(-3, 3, 500).map((v) => {
      return {
        x: v,
        y: dist.dt(v, 7)
      };
    }),
    t_dist5: d3.ticks(-3, 3, 500).map((v) => {
      return {
        x: v,
        y: dist.dt(v, 5)
      };
    }),

    uniform: d3.ticks(-2.1, 2.1, 500).map((v) => {
      return {
        x: v,
        y: dist.dunif(v, -2, 2)
      };
    })
  };
}
```

```{ojs}
fill_limits = (mult) => {
  let s = sd_value * mult;
  return [
    { x: mean_value - s > -4 ? mean_value - s : -4, y: 0 },
    normal_plot(
      mean_value - s > -4 ? mean_value - s : -4,
      mean_value - s > -4 ? mean_value - s : -4,
      mean_value,
      sd_value
    )[0],
    normal_plot(
      mean_value + s < 4 ? mean_value + s : 4,
      mean_value + s < 4 ? mean_value + s : 4,
      mean_value,
      sd_value
    )[0],
    { x: mean_value + s < 4 ? mean_value + s : 4, y: 0 }
  ];
}
```

```{ojs}
function dnorm(x, mean, sd) {
  return jstat.normal.pdf(x, mean, sd);
}
```


