```{r}
#| echo: false
#| include: false

require(ggplot2) 
require(cowplot)
require(tidyr)
require(dplyr)
require(plotly)
salary <- readr::read_csv(here::here("data","world_salary.csv"))
USD <- scales::dollar_format(suffix = " USD")
ggplot2::theme_set(cowplot::theme_cowplot())
```

In Lecture 4 we started talking about how quantitative methods deal with
**measurement**, and we started talking about putting numbers to things.
In today's lecture, we're going to start talking about what we actually do with
these numbers, and how to make a little more sense of measurements in general.
This lecture will be the first in a set of lectures where we'll talk about
**samples** and **populations**, how to **describe** them, and how to understand
the relationship between them. A lot of these ideas are interconnected, so we'll
be touching on some ideas multiple times. But hopefully each time we'll be able
to gain a richer understanding.

The first thing we'll want to do when we've collected a set of measurements is
to describe them in some way. One way to do this is to work out what the
**typical value** is. And it's this kind of description that we'll turn our
attention to first. We'll start by looking at three different ways we can
describe the **typical value** in a set of numbers. 

## Measures of central tendency

What we mean by the **typical value** is not always clear. For example, in
@fig-income we can see the average annual salary (in US dollars) for a set of
`r dim(salary)[1]` countries. Each bar of the plot represents the number of
countries in the given salary bracket ($0--$10k, $10-$20k, …).

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig.align: center
#| fig.height: 3
#| label: fig-income
#| fig.cap: |
#|  National average annual salary [source: worlddata.info]

(salary |> ggplot(aes(x = yearly)) +
  geom_histogram(binwidth = 1e4, boundary = 0, fill = "white", color = "black") +
  labs(x = "Annual salary in USD", y = "Count")) |>
  plotly::ggplotly(tooltip = "count") |>
  plotly::config(displayModeBar = F)
```


From the plot, we can see that there are a lot of countries where the average
annual salary is less than USD 30,000. There are also a handful of countries
where the average annual salary is more than USD 100,000. What would you
consider the _most typical_ annual salary? The bracket with the most countries
in it? If so, that would mean the most typical salary is between $0 and USD
10,000 per year. Or maybe we should pick the value where half the countries
have a lower average salary and half the countries have a higher one? Choosing
this option leads to an estimate of the most typical salary of `r salary$yearly |> median() |> USD()` per year.

As you can see, depending on how we define _the most typical value_, we get
different answers. We'll cover the three main ways of defining the _typical
value_ or _average_. Together, these ways of describing the _typical_ or
_average_ value are known as **measures of central tendency**.

### Mode

The **mode** is a term that refers to the **most frequent value** in a set of
measurements. This is the kind of _average_ we discussed above when we said the
most typical salary is between USD 0 and USD 10,000 a year. The
easiest way to spot the **mode** is to draw a plot like the one we did in
@fig-income and then look for the tallest bar.

A set of numbers can have one or more modes. If it has one mode, then it is
said to be **unimodal**. **Bimodal** means it has two modes. If it has three or
more modes, then it's usually called **multimodal**. Some examples of
this are shown in @fig-diff-modes.

```{r}
#| echo: false
#| label: fig-diff-modes
#| fig-cap: Datasets with different numbers of modes.
#| fig-subcap:
#|    - "A unimodal dataset"
#|    - "A bimodal dataset"
#|    - "A multimodal dataset"
#| layout-ncol: 3
require(ggplot2)
require(dplyr)

# Let's draw some plots

# Unimodal
set.seed(123)
tibble(x = rchisq(n = 10000, df = 5, ncp = 0)) |>
  ggplot(aes(x = x)) +
  geom_histogram(bins = 50, boundary = 0, col = "grey20", fill = "grey80") +
  theme_cowplot() +
  labs(x = "value", y = "count")

# Bimodal
set.seed(123)
d <- rnorm(n = 5000, mean = 2, 1)
tibble(x = c(
  d,
  d + 3
)) |>
  ggplot(aes(x = x)) +
  geom_histogram(bins = 50, boundary = 0, col = "grey20", fill = "grey80") +
theme_cowplot() +
labs(x = "value", y = "count")

# Multimodal
d <- rep(seq(0, 5, length.out = 100), times = 100)
tibble(x = d) |>
  ggplot(aes(x = x)) +
geom_histogram(bins = 11, boundary = 0, col = "grey20", fill = "grey80") +
theme_cowplot() +
labs(x = "value", y = "count")
```


:::{.callout-important}

The mode is the only definition of _typical value_ that works for data that is
measured at the **nominal**/**categorical** level (see Lecture 4).

When it comes to truly **continuous** variables, such as height, the mode is
often not very informative. Why? Take the example of reaction time.
Is it likely that two people will have **exactly** the same reaction time, right
down to the exact number of nanoseconds? Or would each reaction time be unique?
If so, each value in the dataset would be unique and, therefore, each value in
the data set would be the mode. For this reason, the mode is rarely used for
continuous variables measured at the **interval** or **ratio** levels.

Average salary is a continuous variable, but we've turned it into a discrete
variable by placing measurements into discrete ranges. By doing so, we can make
the mode useful.

:::

### Median

```{r}
#| echo: false
dice_res <- c(3, 4, 6, 1, 1) 
serial_and <- function(x) {
  n <- length(x)
  first <- x[1:n-1]
  second <- x[n]
  glue::glue('{paste0(first, collapse = ", ")}, and {second}')
}
```

The **median** is the second kind of average we talked about
[above](#measures-of-central-tendency); the middle value where half the
measurements are above that value and half the measurements are below. To find
the median, we first need to sort our data. Let's say we roll a 6-sided dice 5
times and get the following: `r serial_and(dice_res)`

To calculate the median, we'll do two steps:

1. Sort the data from smallest to largest: `r sort(dice_res)`

2. Find the mid-point: We have five observations so the third one in the
sorted sequence is the mid-point.

Out of the five rolls the median is `r median(dice_res)` (and the mode is 1). If
we had an even number of observations then the median would be the half-way
point between the two mid-point values. For example, if we instead rolled the
dice 6 times and got the results 1, 1, 3, 4, 4, and 6 then the median would be
the midpoint between 3 and 4, or `r median(c(1, 1, 3, 4, 4, 6))`.

@fig-sorted below shows the annual salary in USD for each of the countries in the
dataset, sorted from lowest to highest. Notice, that this time, we’re not
grouping countries into salary brackets and looking at how many there are in
each, as was the case in @fig-income. Here, each bar represents a country.

Because we have an even number of countries in our dataset (78), there are two
mid-points. These are highlighted in orange in @fig-sorted. To get the median
annual national salary, we need to find the value half-way between Romania and
Venezuela, which in this data set turns out to be `r median(salary$yearly) |>
USD()`.


```{r}
#| echo: false
#| label: fig-sorted
#| fig.cap:   Average national salary per country sorted from lowest to highest (Hover over the bars to see the name of the country and the value).
(salary |>
  dplyr::arrange(yearly) |>
  dplyr::mutate(index = 1:n()) |>
  dplyr::mutate(
    color =
      case_when(
        index == (1:78 |> median() |> floor()) ~ "red",
        index == (1:78 |> median() |> ceiling()) ~ "red",
        TRUE ~ " grey"
      )
  ) |>
  dplyr::mutate(text = glue::glue("{country} (${round(yearly/1000,2)}k)")) |>
  ggplot(aes(x = index, y = yearly, text = text, fill = color)) +
  ggplot2::scale_fill_manual(
    name = NULL,
    values = c("#454A60", "#ff4500"), labels = NULL, guide = "none"
  ) +
  labs(y = "Annual salary in USD", x = "Country") +
  scale_x_continuous(labels = NULL, breaks = NULL) +
  scale_y_continuous(labels = function(x) paste0(x / 1000, "k")) +
  ggplot2::theme(
    axis.text.x = element_blank(), axis.ticks.x = element_blank(),
    legend.position = "none", legend.title = element_blank(),
    panel.grid.major.y = element_line(color = "grey60"),
    axis.line.x = element_blank(), axis.line.y = element_blank()
  ) +
  geom_col() +
  NULL
) |>
  plotly::ggplotly(tooltip = "text") |>
  plotly::config(displayModeBar = F)
```


:::{.callout-important}

To be able to calculate a meaningful median, the variable must be measured on
**at least the ordinal level**. For example, if we had categorical data like eye
colour, then it wouldn't make sense to ask for the median between a set of four
blue eyes and 3 green eyes.

:::

### Mean

The arithmetic mean is what most people are referring to when they talk about
the _average_. You can work out the mean of a set of numbers by adding up all the
values and then dividing this by the number of values in the data set.
Mathematically, you can represent this with the formula shown in @eq-mean,
below:


$$\bar{x}=\frac{\displaystyle\sum^{N}_{i=1}{x_i}}{N}$$ {#eq-mean}


This formula just tells use that the mean ($\bar{x}$) is equal to the sum of
all the numbers ($\sum^{N}_{i=1}{x_i}$) divided by the number of values in the
dataset ($N$).

:::{.callout-note}

We use the symbol $\bar{x}$ (pronounced x-bar) to represent the mean of a sample
of data. We use the symbol $\mu$ (pronounced mew) to represent the mean of the
population (see
[sample means and population means](#sample-means-and-population-means) below).

In fact, we usually use Greek letters for population values (parameters) and
Latin letters for sample value (statistics).


:::



### Mean vs Median

Both the mean and median have their advantages and disadvantages. The mean
is easier to work with from a mathematical point of view. And for this reason,
most of the statistical methods we'll be learning about are based on the mean.

Compared to other measures of central tendency, _means taken from different
samples of the same population_ tend to be **more similar to each other**.
(We'll talk more about samples and populations
[below](#sample-means-and-population-means)). If, on the other hand, we
calculated medians for different samples from the same population, there would
be more variability in the values we'd get.

Turning back to our dice example:
If we took the median and the mean of 5 dice rolls, and we did this over and
over, the mean values would be more bunched around 3.5 with very few values less
than 2. The median values would still be centred around 3.5, but we'd get
more values less than 2, so they would be more spread out.

There are also some downsides to the **mean** relative to the
**median**. The primary one is that the _mean_ is **sensitive to extreme
values** in a way that the _median_ is not. This means that, even if we have a
really big sample size, adding a **single value** that is extremely big or
extremely small can shift the _mean_ dramatically. This is not the case for
the _median_.


```{r}
#| echo: false

d1 <- c(5, 3, 1, 7, 10, 4, 5)
n_v <- 1000
d2 <- c(d1, n_v)
```


For example, let's say we have a list of numbers: `r d1`. This list of numbers
has a median of `r median(d1)` and a mean of `r mean(d1)`. If we added a
value of `r n_v` to this list, then our new median would be unchanged
at `r median(d2)`, but our new mean would be `r mean(d2)`. You can explore the
influence of data points on the mean and the median in @exm-mean-median.


::::{.callout-tip icon="false" appearance="simple"}

:::{#exm-mean-median}
#### Explore the mean and median
:::

Click on the plot below to add data to the data set. After you've placed a data
point, you can click and drag it to move it. You can use the check boxes
to display the mean (red) and median (blue).

To see how the <font color="red">**mean**</font> and the <font
color="blue">**median**</font> are influenced by extreme values try adding
several points to the left side of the plot before adding a single point to the
far right. Drag this point around moving it closer or further away from the
left side points. Notice how the <font color="blue">median</font> is unchanged
but the <font color="red">mean</font> does change.

Tick the *show data table* box to show the raw data. From this you should be
able to work out the mean and median yourself. You can try doing this to make
sure you understand how they're calculated.

::: {.content-hidden when-format="pdf"}

```{ojs}
//| echo: false
import {central_tendency_display} from "@ljcolling/measures-of-central-tendency"
import {viewof ave_select} from "@ljcolling/measures-of-central-tendency"
import {viewof clear} from "@ljcolling/measures-of-central-tendency"
import {viewof show_data_table} from "@ljcolling/measures-of-central-tendency"
import { data_table} from "@ljcolling/measures-of-central-tendency"
import {viewof summaryText} from "@ljcolling/measures-of-central-tendency"
central_tendency_display
```

\
```{ojs}
//| echo: false
//| panel: input
viewof ave_select
viewof clear
viewof show_data_table
```

```{ojs}
//| echo: false
data_table
```

```{ojs}
//| echo: false
viewof summaryText
```

:::

::: {.content-hidden when-format="html"}

**Not available in PDF handout**

**View the online version of the handout**

:::

::::



## Sample means and population means

So far we've talked about describing the typical value in a set of measurements
that we have---our **sample**. But one of the key things that we want to do
with *statistics* is to make **inferences** about **populations** from *the
information* that we get from **samples**. That is, we often want to make a
judgement, or draw a conclusion, about an aspect of the population when all we
have access to is a sample. We'll get to more formal definitions of
_populations_ and _samples_ shortly, but first, let's make things more concrete
by introducing an example.

Let's say you're interested in the **average height** of **people in the UK**.
The "easy" way to find an answer to this question is to measure **all the people
in the UK** and then work out the **average height**. Doing this will give you
the exact answer to your question. But if you can't measure everyone in the UK,
then what do you do?

You could instead select a smaller group, or subset, of people from the UK. You
could then measure the height of people in this group, and then try to use this
information to figure out plausible values for the average height of people in
the UK.

In this example, the group (or groups) you're making claims about is the
population---you want to make claims about **the average height** of **people
in the UK**. And the **sample** is a subset of this population---the smaller
group of people that you were eventually able to measure.

It's important to note that there isn't a **single** population. What counts as
the population will depend on the claim you're making. For example, let's say
I'm interested in testing the claim, "Do **people in East Sussex** show an
interference effect on the Stroop task?". Here the **population** would be
**people in East Sussex**. If, however, I want to make claims about **people in
general**, then the **population** might be all **living** humans.

### Theoretical populations

We often talk about populations as if they're a **set of actually existing
things** that we can take our sample from---for example, all **living** humans.
But populations don't have to be sets of actually existing things. Instead, they
can be the **set of possible things** from which our samples can be drawn. This
is a fairly advanced idea, so it might seem a little confusing for now. But
keeping this idea in the back of your mind will help statistics make more sense
as you progress through your degree. Because this idea is a little confusing, an
example might help.


<!-- FIXME: This should be switched to dice rolls -->

Let's say we want to collect a sample of 2 dice rolls. To collect our
**sample** we take a die and roll it twice. From this we can easily
work out the typical value—that is, the _mean_—just as we did above.

Our **sample** is the set of 2 dice rolls that we've collected. But what is our
**population**? One way to think of our population is as the **set of possible
outcomes that could occur if we rolled a die twice**. What is the typical
value of this population? It turns out that we can actually work this out.

To work it out we would do something like the following:

::: {.content-hidden when-format="pdf"}

```{ojs}
//| echo: false
import {viewof view} from "@ljcolling/probability"
viewof view
```

:::

::: {.content-hidden when-format="html"}

**Not available in PDF handout**

**View the online version of the handout**

:::

This diagram is a little complex, so we'll step through it. Each colour
represents rolling a different number. So, for example, the blue circle
means that you rolled 1, the green is rolling a 2, the yellow is rolling
a 3, black for 4, white for 6, and purple for 6.

The diagram shows all the possible things that could happen if we rolled the 
dice twice. So for example, one thing that could happen is we could roll a 
1 (blue) followed by another 1 (blue). Or we could roll a 1 (blue) followed
by a 2 (green). 

So how do we work out the typical value? First we work out how many of the 
sequences sum up to 2 (the minimum total you can get with two dice rolls). 
We then count up the number of sequences that sum up to 3, to 4, 5, and so
on all the way up to 12 (the maximum total you can get with two dice rolls).

:::{.callout-note collapse="true"}

#### Expand to see all the possible outcomes

```{r}
#| echo: False
tibble::tibble(roll1 = 1:6, roll2 = 1:6) |> 
  as.list() |> 
  purrr::cross_df() |>
  mutate(total = roll1 + roll2) |>
  knitr::kable(col.names = c("1st Roll", "2nd Roll", "Total"))


```
:::

After we've done this we'll see that more sequences add up to 7 than any other
total. So we could say that the typical outcome is an outcome that leads to the
two dice adding up to 7. If we had a set of 2 dice rolls where the sum was 7,
then the mean would be 3.5 (because 7 ÷ 2 = 3.5). So the **population mean**
for two dice rolls is 3.5.

We'll talk more about this idea when we start talking about distributions, but
for now this basic description should suffice.

<!-- END OF FIXME: -->

We can work it out because we know something about the **process that gives rise
to our data** (We'll talk more about how we know this in a later lecture on
probability). That is, we can characterise the process that gives rise to the
samples, and our samples are just a set of instances of data generated by this
process.

Taking this idea and applying it to the Stroop task, where we're interested in
**humans in general**, we might want to say that our population isn't
just all living humans. Instead, we might want to say that our population is all
possible humans that might have lived, might be living now, and will live in the
future. Or that our samples are just some instances of data generated by some
process that goes on in people's brains when they do the Stroop task. Unlike the
dice roll example, we don't know the characteristics of this process. This means
that we couldn't, for example, draw out what the population would look like. But
the point of doing science and statistics is so that we can use the information
from samples to start to characterise this process and start to work out what
the population might look like. But how do we go from samples to populations?

### The relationship between samples and populations

Let's assume that we have explicitly defined our **population** as
_all people in the UK_, and we've collected a **sample** by taking measurements
from a **subset** of this population. What is the relationship between this
sample and the population from which it was drawn?

The _sample_ should **resemble** the _population_ in some way. Most often we're
interested in knowing something like: "_What is the typical value (i.e., the
mean) of the population?"_ So for our height example, this would mean that we're
interested in knowing the **mean height of the population**. Ideally then, the
**average height** of our **sample** should **resemble** the **average** height
of our **population**. But if we don't know the **average** of our **population**,
then how will we know whether our **sample** _resembles_ it?


For purposes of illustration, let's say that I actually know what the
population looks like. We'll say that adults in the UK range in height between
78cm and 231cm, but that the average height of an adult in the UK is 170cm. So
our population mean is 170cm. Now let's collect a sample of data. We'll talk
more about the influence of sample size in a later lecture, but for now let's
just say that we collect a sample of 50 people. And let's say that we don't
only collect one sample, but that we collect a sample of 50 people over and
over again. In @exm-mean-samples you can see the average height of each of our
samples of 50 people, with the solid line showing the population mean.


::::{.callout-tip icon="false" appearance="simple"}

:::{#exm-mean-samples}
#### Explore sample means
:::

In the plot below, each dot represents the mean height from a sample of 50
people from our simulated UK population. The solid horizontal line shows the
actual population mean of our simulated UK population. The plot shows the means
for 100 different samples of 50 people. Press **Replay** to generate a fresh set
of 100 samples of 50 people.


::: {.content-hidden when-format="pdf"}

```{ojs}
//| echo: false
import {sample_means} from "@ljcolling/measures-of-central-tendency"
import {viewof replay_mean} from "@ljcolling/measures-of-central-tendency"
sample_means()
```

\
```{ojs}
//| echo: false
//| panel: input
viewof replay_mean
```



:::

::: {.content-hidden when-format="html"}

**Not available in PDF handout**

**View the online version of the handout**

:::

::::

What do you notice in @exm-mean-samples? The sample means don't always line up
exactly with the population mean. Sometimes the sample mean is higher and
sometimes the sample mean is lower. It moves around a bit from sample to
sample. Because it moves around, and because we don't know the population mean,
this tells us that **on any particular sample** we won't know whether the
sample mean is the same as the population mean.


### The average of the sample means

But let's think of things from a slightly different perspective. Let's treat
the mean of each sample of 50 people as a measurement. We'll now take a
**"sample"** of these measurements. That is, we'll measure the height of 50
people and we'll work out the average height. This might be something like 168
cm. We'll then measure another 50 people, and we'll work out the average height,
which might be something like 175 cm. We'll then work out the average of these
two averages (`r (168 + 175) / 2` cm). We'll do this over and over recalculating
our average of averages after each new sample of 50 people. In
@exm-running-samples we can see what happens to our average of our
**sample of samples**. What do you notice?


::::{.callout-tip icon="false" appearance="simple"}

:::{#exm-running-samples}
#### Explore sample of samples
:::

In the plot below we can see the average of our sample of samples (or our
_running average_) as we add more samples. That is, we can see what the average
would be after 2, 10, or 100 samples (etc), where at each step we're adding the
average of our new sample to the set of averages of all the samples that have
come before. Click **replay** to start the process over with a new set of
samples.

::: {.content-hidden when-format="pdf"}
```{ojs}
//| echo: false
import {sample_means_ave} from "@ljcolling/measures-of-central-tendency"
import {viewof replay_running_mean} from "@ljcolling/measures-of-central-tendency"
sample_means_ave()
```

\
```{ojs}
//| echo: false
//| panel: input
viewof replay_running_mean
```
:::

::: {.content-hidden when-format="html"}

**Not available in PDF handout**

**View the online version of the handout**

:::

::::

That's right, as we carry on collecting more and more **samples of samples** the
average of these will eventually line up and match the **population mean**. This
shouldn't be completely surprising. If we collect enough samples, and combine
them all together, then eventually we'll have sampled the entire population.

But the point I want you to take away from this demonstration is the following:
This tells us that even though we don't know whether the mean of **any
particular sample** is the same as the population mean, the **sample mean**
will **on average** be the same as the population mean. We'll touch on this
idea more in future lectures when we talk about the **sampling distribution**,
but for now this simple idea is all you need to know. 

## Check your understanding

Use this quiz to make sure that you've understood the key concepts.

```{r}
#| echo: false
#| message: false

require(niceQuiz)

quiz(
  caption = "",
  question(
    "What is the only measure of central tendency that can have multiple values in a single sample?",
  answer("Mean"),
  answer("Mode", correct = TRUE),
  answer("Median"))
)


quiz(
  caption = "",
  question(
    "Here is a set of numbers: </br> 34, 5, 42, -18, 4, 5<br />What is the median of these numbers?",
    numericanswer(c(34, 5, 42, -18, 4, 5) |> median()),
    type = "fillin"),
  question(
    "What is the mode of the same set of numbers?",
    numericanswer(5),
    type = "fillin"),
  question(
    "What is the mean of the same set of numbers?",
    numericanswer(c(34, 5, 42, -18, 4, 5) |> mean()),
    type = "fillin")
)


quiz(
  caption = "",
  question(
    "Which of these measures will change the most if we add -577 to our numbers and then recalculate? <br /> Try to answer without recalculating the numbers.",
  answer("Mode"),
  answer("Impossible to tell"),
  answer("Mode"),
  answer("Mean", correct = TRUE)
))

```

