```{r}
#| echo: false
#| include: false

require(ggplot2) 
require(cowplot)
require(tidyr)
require(dplyr)
require(plotly)
salary <- readr::read_csv(here::here("data","world_salary.csv"))
USD <- scales::dollar_format(suffix = " USD")
ggplot2::theme_set(cowplot::theme_cowplot())
```

## Today's lecture

- So far we've talked about quantitative methods in the abstract

- We've said quantitative methods is all about putting **numbers to things**, but
  we haven't talked about what to do with the numbers

In this lecture, and the one that follows we're going to start talking
about the techniques we can use for **describing sets of measurements**

We'll use these tools when we start learning the basics of **statistical
models**, and learn about **the sampling distribution**

---

## Measures of central tendency

When we have a set of measurements the first thing we might want to do is
work out what the typical value is


```{r}
#| echo: false
#| warning: false
#| message: false
#| fig.align: center
#| fig.height: 3
#| label: fig-income
#| fig.cap: |
#|  National average annual salary [source: worlddata.info]

(salary |> ggplot(aes(x = yearly)) +
  geom_histogram(binwidth = 1e4, boundary = 0, fill = "white", color = "black") +
  labs(x = "Annual salary in USD", y = "Count")) |>
  plotly::ggplotly(tooltip = "count") |>
  plotly::config(displayModeBar = F)
```

What we mean by **typical value** is not always clear

- What is the _typical_ average income in the set of countries shown in
  @fig-income?

- Lot's of countries have an income below USD 30,000, but some have
  incomes higher than USD 100,000

---

## Measures of central tendency

_Maybe we should pick the bracket with the most countries in it?_

- Then the typical salary is between $0 and USD 10,000 per year. 

_Maybe we should pick the value where half the countries have a lower
  average salary and half the countries have a higher one?_

- Then the typical salary is `r salary$yearly |> median() |> USD()` per
year.


Depending on how we define _the most typical value_, we get
different answers. 

We'll cover the three main ways of defining the _typical value_ or
_average_

Together, these ways of describing the _typical_ or _average_
value are known as **measures of central tendency**.

---

### Mode

 - The **mode** is the **most frequent value** in a set of
measurements

- For this kind of _average_ the most typical salary is between $0 and $10,000 

- The easiest way to spot the **mode** is to draw a plot like the one we did in
  @fig-income and then look for the tallest bar

A set of numbers can have more than one mode. Some examples of this are shown
in @fig-diff-modes.

```{r}
#| echo: false
#| label: fig-diff-modes
#| fig-cap: Datasets with different numbers of modes.
#| fig-subcap:
#|    - "A unimodal dataset"
#|    - "A bimodal dataset"
#|    - "A multimodal dataset"
#| layout-ncol: 3
require(ggplot2)
require(dplyr)

# Let's draw some plots

# Unimodal
set.seed(123)
tibble(x = rchisq(n = 10000, df = 5, ncp = 0)) |>
  ggplot(aes(x = x)) +
  geom_histogram(bins = 50, boundary = 0, col = "grey20", fill = "grey80") +
  theme_cowplot() +
  labs(x = "value", y = "count")

# Bimodal
set.seed(123)
d <- rnorm(n = 5000, mean = 2, 1)
tibble(x = c(
  d,
  d + 3
)) |>
  ggplot(aes(x = x)) +
  geom_histogram(bins = 50, boundary = 0, col = "grey20", fill = "grey80") +
theme_cowplot() +
labs(x = "value", y = "count")

# Multimodal
d <- rep(seq(0, 5, length.out = 100), times = 100)
tibble(x = d) |>
  ggplot(aes(x = x)) +
geom_histogram(bins = 11, boundary = 0, col = "grey20", fill = "grey80") +
theme_cowplot() +
labs(x = "value", y = "count")
```

---

### Mode

- The mode is the only definition of _typical value_ that works for data that is
measured at the **nominal**/**categorical** level (see Lecture 4).

- When it comes to truly **continuous** variables, such as height, the mode is
often not very informative. Why?

  - Because each value in the dataset is probably unique

  - For this reason, the mode is rarely used for continuous variables measured
    at the **interval** or **ratio** levels


Average salary is a continuous variable, but we've turned it into a discrete
variable by placing measurements into discrete ranges. By doing so, we can make
the mode useful.

---

### Median

```{r}
#| echo: false
dice_res <- c(3, 4, 6, 1, 1) 
serial_and <- function(x) {
  n <- length(x)
  first <- x[1:n-1]
  second <- x[n]
  glue::glue('{paste0(first, collapse = ", ")}, and {second}')
}
```

- The **median** is the middle value where half the measurements are above that
  value and half the measurements are below

- The easiest way to work out the media is to sort our data (e.g., from
  the smallest to the largest)


Let's say we roll a 6-sided dice 5 times and get the following: `r serial_and(dice_res)`

To calculate the median, we'll do two steps:

  1. Sort the data from smallest to largest: `r sort(dice_res)`

  2. Find the mid-point: We have five observations so the third one in the
  sorted sequence is the mid-point.

Out of the five rolls the median is `r median(dice_res)` (and the mode is 1).

---

### Median

```{r}
#| echo: false
#| fig.align: center
#| out.height: 65%
#| out.width: 100%
#| label: fig-sorted
#| fig.cap:   Average national salary per country sorted from lowest to highest (Hover over the bars to see the name of the country and the value).
(salary |>
  dplyr::arrange(yearly) |>
  dplyr::mutate(index = 1:n()) |>
  dplyr::mutate(
    color =
      case_when(
        index == (1:78 |> median() |> floor()) ~ "red",
        index == (1:78 |> median() |> ceiling()) ~ "red",
        TRUE ~ " grey"
      )
  ) |>
  dplyr::mutate(text = glue::glue("{country} (${round(yearly/1000,2)}k)")) |>
  ggplot(aes(x = index, y = yearly, text = text, fill = color)) +
  ggplot2::scale_fill_manual(
    name = NULL,
    values = c("#454A60", "#ff4500"), labels = NULL, guide = "none"
  ) +
  labs(y = "Annual salary in USD", x = "Country") +
  scale_x_continuous(labels = NULL, breaks = NULL) +
  scale_y_continuous(labels = function(x) paste0(x / 1000, "k")) +
  ggplot2::theme(
    axis.text.x = element_blank(), axis.ticks.x = element_blank(),
    legend.position = "none", legend.title = element_blank(),
    panel.grid.major.y = element_line(color = "grey60"),
    axis.line.x = element_blank(), axis.line.y = element_blank()
  ) +
  geom_col() +
  NULL
) |>
  plotly::ggplotly(tooltip = "text") |>
  plotly::config(displayModeBar = F)
```

Because we have an even number of countries in our dataset (78), there are two
mid-points. 

To get the median annual national salary, we need to find the value half-way
between Romania and Venezuela, which is `r median(salary$yearly) |> USD()`.

---

### Median


- To be able to calculate a meaningful median, the variable must be measured on
  **at least the ordinal level**. 

- If we had categorical data like eye colour, then it wouldn't make sense to
  ask for the median between a set of four blue eyes and 3 green eyes

---

### Mean

- The  mean is what most people think of when we talk about the _average_

- You can work out the mean by **adding up all the values** and then diving
  this by the number of values


Mathematically, you can represent this with the formula shown in @eq-mean,
below:

$$\bar{x}=\frac{\displaystyle\sum^{N}_{i=1}{x_i}}{N}$$ {#eq-mean}


You can also write @eq-mean as @eq-mean-sum

$$\bar{x}={\displaystyle\sum^{N}_{i=1}{\frac{x_i}{N}}}$$ {#eq-mean-sum}

@eq-mean might be more familiar to you, but I like @eq-mean-sum because it
makes it clear that a **mean** is just a special way of adding up numbers (more
in Lecture 8)

---

### Mean


$$\bar{x}=\frac{\displaystyle\sum^{N}_{i=1}{x_i}}{N}$$ {#eq-mean}


- This formula just tells use that the mean ($\bar{x}$) is equal to the sum of
all the numbers ($\sum^{N}_{i=1}{x_i}$) divided by the number of values in the
dataset ($N$).


- We use the symbol $\bar{x}$ (pronounced x-bar) to represent the mean of a sample
of data. 

- We use the symbol $\mu$ (pronounced mew) to represent the mean of the population

---

### Mean vs Median


Both the mean and median have their advantages and disadvantages

- The mean is easier to work with from a mathematical point of view

- Means taken from different samples of the _same population_ tend to be
  **more similar to each other** (see @fig-dice)

```{r}
#| echo: false
#| fig.cap: Means and medians for 5 dice rolls repeated 10,000 times
#| label: fig-dice
#| out.height: 60%
#| out.width: 50%
# require(tidyverse)
# require(plotshow)

set.seed(123)
gen <- function(x) {
  sample.int(n = 6, size = 5, replace = TRUE) |>
    pipebind::bind(b, data.frame(Mean = mean(b), Median = median(b)))
}
df <- parallel::mclapply(X = seq_len(10000), gen) |>
  bind_rows() |>
  tidyr::pivot_longer(1:2)

df |>
  ggplot(aes(x = value, group = name)) +
  geom_histogram(colour = "white", fill = "black") +
  facet_grid(~name) +
  labs(y = "Number of samples", x = "Value") +
  scale_x_continuous(breaks = seq_len(6)) +
  theme_minimal() +
  NULL
```

---

### Mean vs Median

There are also some downsides to the <font color="red">mean</font> relative
to the <font color="blue">median</font>.

The <font color="red">mean</font> is **sensitive to extreme values** in a way
the <font color="blue">median</font> is not

```{r}
#| echo: false
require(htmltools)
vid <- tags$video(width = "100%", autoplay = NA, loop = NA, controls = NA)
source <- tags$source(src = here::here("images","test.webm"), type = "video/webm")
vid$children = source
vid
```

---


## Sample means and population means


- So far we've just talked about describing the typical value in a set of
  measurements that we have---our **sample**

- But we want to do with *statistics* is to make **inferences** about
  **populations** from *the information* that we get from **samples**

- If you're interested in the **average height** of **people in the UK** the
  "easy" way to find an answer to this question is to measure **all the people
  in the UK** and then work out the **average height**

_But if you can't measure everyone in the UK, then what do you do?_

- You could instead select a smaller group, or subset, of people from the UK.
  Measure the height of people in this group, and then try to use this
  information to figure out plausible values for the average height of people
  in the UK.

In this example, the group (or groups) you're making claims about is the
**population**, and the **sample** is a subset of this population

---

### Theoretical populations

- We often talk about populations as if they're a **set of actually existing
  things** that we can take our sample from---for example, all **living**
  humans

- But populations don't have to be sets of actually existing things. Instead,
  they can be the **set of possible things** from which our samples can be
  drawn

- Let's say we want to collect a sample of 2 dice rolls

  - To collect our **sample**, we take a die and roll it **twice**

  - We can then work out the typical value (i.e., the mean) from these rolls

Our **sample** is the set of 2 dice rolls we've collected, but what is our
**population**?

One way to think of our population is as the **set of possible outcomes that
could occur if we rolled the dice twice**


---

### Theoretical populations

If our population is **all possible rolls of two dice** then what is the mean
of this population?

We can easily draw out all the possible things that will happen if we roll a
dice twice:





From this, we can count up how many times we get a total 2, 3, 4, etc from two
dice rolls

- We'd find that 6 sequences lead to a total of 7 (more than any other total(

- A total of 7 gives a mean of 3.5 (7 รท 2 = 3.5)

---

### Theoretical populations

- We can work out the **population mean** of two dice rolls because we know
  something about the data generating process 

- Our samples are just a set of instances of data generated by this process


- Applying this idea to something like the Stroop task we say that:

    - Our population isn't **all living humans** but all humans that might have
      lived, might be living now, and might be living in the future

    - Our samples are just instances of data generated by the process that goes
      on in peoples brains when they do the Stroop task

For the Stroop task (or any other psychological process we might be interested
in) we can't just work out exactly what that data generating process looks like

So we collect samples to try to characterise it

---

## From samples to populations

Let's say we have defined our **population** as _all people in the UK_

Our **sample** is a **subset** of this

- We really want to know about the **population**. E.g., What is the _average_
(**mean**) height of _people in the UK_?

- But all we have is our **sample**. I.e., The _average_ (**mean**) height of
_people in our sample_.

If we want to go from our **sample** to the **population** then _ideally_ our
**sample mean** should **resemble** our **population mean**

But if _real life situations_ we don't know the **population mean**, so how
would we know whether our **sample mean** **resembles** it?

---

### A sample of samples

- Let's say I happen to know that the population mean for the height of people
  in the UK is 170cm (and heights range from about 78cm to 231 cm)

- I can now draw a sample of 50 people from this population and calculate the
  mean 

- I can do this over and over and plot each sample mean in @fig-samps

```{ojs}
//| echo: false
import {sample_means} from "@ljcolling/measures-of-central-tendency"
import {viewof replay_mean} from "@ljcolling/measures-of-central-tendency"
```

```{ojs}
//| fig-cap: Repeatably sampling our population
//| label: fig-samps
//| echo: false
sample_means(1200, 300)
```

```{ojs}
//| echo: false
viewof replay_mean
```

---

### A sample of samples

- The samples don't always line up exactly with the population mean

- Sometimes it's higher, sometimes it's lower. Sometimes it's closer and
  sometimes it's further way

- Because we **don't know the population** we'd never know whether **any
  particular sample** was close, far, higher, or lower **than the population
  mean**

Even though we can't say that **a particular sample** is close to the
population, there is something else we **can** say

We can say how **sample means** will behave **on average**:


> The sample mean will **on average** be the same as the population mean


\
_This is an idea that we'll make use of **a lot** in statistics_

_But what does it **mean**?_

---

### The average of the sample means

- If treat each sample mean from 50 people as a **measurement** 

- As we collect more samples, we average together the **sample means**


```{ojs}
//| echo: false
import {sample_means_ave} from "@ljcolling/measures-of-central-tendency"
import {viewof replay_running_mean} from "@ljcolling/measures-of-central-tendency"
```

```{ojs}
//| fig-cap: Running average of sample means
//| label: fig-running
//| echo: false
sample_means_ave(1200, 300)
```

```{ojs}
//| echo: false
viewof replay_running_mean
```

This **average** of **sample means** will eventually be the same as the
**population mean**<sup>1</sup>

<aside><sup>1</sup>In technical terms we say: $\mathrm{E}(\bar{x}) = \mu$</aside>

---

## Looking forward

- So far we've covered **measures of central tendency** for samples

- And we've covered the idea of the **sample mean** ($\bar{x}$) and
  **population mean** ($\mu$)

- We saw that although we don't know whether **a particular sample mean** is
  the same as the **population mean**, we do know that **on average** they will
  be the same

- In the coming lectures we'll learn how to describe how **spread out** our
  **sample** is, and how **spread out** the **population is**


When we plotted the **individual sample means** we saw that they were **spread
out around the population mean**

We'll finally put ideas about **means** and **spread** together to finally work
out how to quantify this spread

_But all that is for later..._
