<style>

kbd {
	margin: 0px 0.1em;
	padding: 0.1em 0.6em;
	border-radius: 3px;
	border: 1px solid rgb(204, 204, 204);
	color: rgb(51, 51, 51);
	line-height: 1.4;
	font-family: Arial,Helvetica,sans-serif;
	font-size: 10px;
	display: inline-block;
	box-shadow: 0px 1px 0px rgba(0,0,0,0.2), inset 0px 0px 0px 2px #ffffff;
	background-color: rgb(247, 247, 247);
	-moz-box-shadow: 0 1px 0px rgba(0, 0, 0, 0.2), 0 0 0 2px #ffffff inset;
	-webkit-box-shadow: 0 1px 0px rgba(0, 0, 0, 0.2), 0 0 0 2px #ffffff inset;
	-moz-border-radius: 3px;
	-webkit-border-radius: 3px;
	text-shadow: 0 1px 0 #fff;
}
</style>

```{r}
#| echo: false

words <- system("pandoc --lua-filter wordcount.lua _content.qmd --from=markdown", intern = TRUE) |> as.numeric()
minutes <- words %/% 200
seconds <- ceiling(((words / 200) - minutes) * 60)
reading_time <- paste0(minutes, " minutes and ", seconds, " seconds")
``` 

`r words` words. Reading time: `r reading_time`

---

## Welcome to Psychology as a Science

The aim of today's lecture is to provide you with an overview of the structure
of the research methods modules that you'll be taking in your degree. I'll also
try to give you some tips that will help you to succeed in this course. Although
this is not an inherently difficult course a lot of the material you'll be
encountering may be new to you. Some of it might even seem a little intimidating
(especially the programming stuff that you'll be doing in the practical
classes). But if you stick with it, and give it your best shot, then I'm
confident that you'll be able to do well.

### Introductions

The teaching team for Psychology as a Science (PAAS) is made up of three people.
Each of us has different roles and responsibilities. Knowing this will help you
to know who to contact if you need help with something (but see more on
[knowing who to email](#knowing-who-to-email) below).

**Dr Lincoln Colling** (lecture content convenor)

Dr Colling is responsible for the lecture content, and for preparing the final
exam.

**Dr Jenny Terry** (practical content convenor)

Dr Terry is responsible for the content covered in the practical sessions, the
`R` homework activities, and the weekly `R` assessments. She is also responsible
of the lab report, the major piece of written work you'll produce this year.[If
you don't know what `R` is then don't worry---99% of the people taking this
course won't know what `R` is. But you'll learn all about `R` and `RStudio` in
the practical classes.]{.aside}

**Dr Vlad Costin**

Dr Costin, along with Dr Terry, is one of the lead practical lecturers.

## Tips for doing well

The information that most of you will want to know is how to do well in this
course. So we'll get it out of the way right up the front. The single best thing
you can do to do well in this course is simply to _show up to class every week_.
Turning up to _in person_ every week not only means that you'll get exposed to
the lecture content, but you'll also get to meet me, talk to me, meet your
classmates, and talk to them too. So much of what makes university special are
those moments that happen around the classes. And all you need to do to
experience those moments is to show up to class every week.

The second most important thing you can do is to do _all the assessments_. This
is particularly important for the weekly assessments linked to the practical
content. Each of these assessments builds on the one before, so if you fall
behind it can be hard to catch up. Even if you find them difficult you should
still do them. They will get easier!

Start off by being present for everything and the rest will fall into place!

## Finding the information you need

The most important thing you need to know if you want to succeed in the course
(and this is true of all the courses you'll be taking) is how to find the
information you need.

The first place to look for any information is on Canvas. If you're reading this
lecture handout then you've probably already found Canvas. The Canvas page for
this course will be your one-stop shop for almost all the information you need
about this course. Familiarise yourself with it. Especially the front page!
Click all the links, and read through the pages. If you have a question like
"What do I do if I can't make a practical or lecture?", then that information is
on Canvas. If you want to know what the various assessments are for this course,
then that will be on Canvas. If you want to know whether references will be
counted as part of the word count for your report, then that will be on Canvas
too!

There's a lot of information on Canvas. In fact, there's probably too much
information on Canvas, and so it can be a little overwhelming. To help with
this, I have two tips.

First, as I said above, click the links and read through all the pages. But
while you're doing this, keep a bit of paper or a notebook handy. Use this to
note down bits of information, or where to find certain information so you can
check back when you need it. You can also bookmark pages that you think will be
useful later. And when you bookmark a page you can give it a informative title
that is meaningful to you. All this will help you to find the information you
need when you need it. And remember, on any web-page you can hit <kbd>Ctrl</kbd>

- <kbd>F</kbd> (windows) <kbd>Command</kbd>
- <kbd>F</kbd> (mac) to search for a word, and this might be particularly useful
  if it's a long page with a lot of text and you're just looking for something
  specific.

Second, check out the [Frequently Asked Questions (FAQ)]({{< var faq >}}). Here
I will list the answers to some questions that I tend to encounter often. You
might just find the answer to your question here. I will try to update this as I
encounter more frequently asked questions, so make sure to check back whenever
you're stuck. And if you have a question that you think your class mates might
also want to know the answer to then click [here]() (this link is also on the
FAQ page) to submit a new Frequently Asked Question, and I'll add it to the
list.

:::{.callout-note}

### Emailing me for information

You might be tempted to just email me as soon as you get stuck for some
information. But remember, there's nearly 600 of you, and I get a lot of email,
so it might take a few days before I can get back to you. So double check the
[FAQ]({{< var faq >}}) and [Canvas]({{< var canvas >}}) before you send that
email, because you might be able to get an instant answer there.

:::

### Knowing who to contact {#knowing-who-to-email}

One thing that is tricky when you're first starting out at university is knowing
who to contact with your queries. Here are a few tips for helping your to figure
out who to contact.

#### Admin queries

Admin queries are any queries than are not directly related to the course
content. For example, if you've been absent from class and you want to inform
somebody, then that's an admin query. If you're meant to have an extension to
hand-in date for an assessment, but it hasn't been correctly applied, then
that's an admin query. If you're unable to submit and assessment because you're
unwell, then that too is an admin query. All admin queries should be directed at
the admin staff.

You can contact the admin staff by emailing them at [](). If you'd specifically
like to report an absence then you can email Psychology Absences at [](). For
more information about Attendance and Absences, check out the
[page on Canvas]().

:::{.callout-important}

Lecturers can't grant extensions for assessments. Please don't email us to ask,
because there's nothing we can do about it.

For information about _Requests for Extended Deadlines_ see the information on
the
[Student Support Unit](https://www.sussex.ac.uk/studentsupport/support-needs/extensionrequests)
page.

:::

#### Queries about course content

The lecturers are you point of content if you have questions about the course
content. Who you contact will depend on what section of the course your question
covers:

**For questions about lecture materials**

If you have a question about material that was covered in one of the regular
weekly lectures then the best person to get in touch with is me (Dr Lincoln
Colling).

I can also help with queries related to the final exam (because the final exam
will only cover the material from the regular weekly lectures).

The best way to get in touch with me is to grab me right after the lecture. I'll
always hang around for a few minutes to answer questions that you might have.
But if you need to run off then you can always book into a drop in session.
There's [information below](#drop-in-sessions) on how to do this.

**For questions about the practical content**

If you have a question about something that was covered in one of your weekly
practical sessions then the best person to contact would be the practical lead
_for your practical_. This will either be Dr Jenny Terry or Dr Vlad Costin.

Your practical lead will also be able to help you out with queries about the
weekly `R` quizzes **For queries about the report**

If you have queries about the report, then the best person to contact is Dr
Jenny Terry.

**For general `R` help**

If you have general `R` and `RStudio` queries, and it not something that need to
be addressed to your practical lead (for example, if it's a general query, and
not, for example, a question about something your lecturer said in class) the
you can go to one of the **R Drop in Sessions**. You _can_ go to any of the R
Drop in Sessions, but try to go to one run by somebody that teaches on PAAS.

To book into an `R`

More information on how to get help with `R` will also be provided in the
practical sessions.

[**Booking a drop in appointment**]{#drop-in-sessions}

Know that you know how to contact, how do you go about actually contacting them?
If it's a question quick question about the lecture content then just come up
and talk to me after class. Or if it's a question about the practical content,
then just ask the lecturer in your practical class. If you question is a little
longer, or if you can't wait until class to ask, then you should book into a
drop in session.

To book into the drop in session you can follow the links on the front page of
the module's canvas page.

## Research methods in Psychology

**Psychology as a Science** is the first of a series of research methods
modules.

Following this, you'll take:

- Analysing Data (next term)

- Discovering Statistics (year 2)

- Quantitative and Qualitative Methods (year 2)

All of these build up to prepare you for the **research dissertation** in your
final year. But they're also a great way to learn a lot of transferable skills
that are useful outside of university, for example:

- How to analyse data

- How to make sense of statistics

- Basic computer programming/coding skills

**Psychology as a Science (PAAS)** covers an introduction to the research
process. Some topics you'll cover include:

- Introduction to Philosophy of Science

- Difference approaches to research including quantitative methods and
  qualitative methods

- Basics of statistical theory

You'll also be introduced to coding in the `R` programming language. You'll
cover the basics of the `R` language, how to process and summarise data, and how
to make plots in `R`.

Along the way we'll work with some of the same data that you'll be using for you
**Cognition in Clinical Contexts** lab report

In **Analysing Data (AD)** you'll learn about the core statistical tests used in
Psychology. You'll also cover moved advanced use of `R`, and you'll learn how to
perform statistical tests in `R`. AD will also give you your first chance to
independently analyse some data.

In **Discovering Statistics (DS)** you'll learn even more advanced statistical
tests, and more in-depth knowledge of `R`.

In the final course, **Quantitative and Qualitative Methods (QQM)** you'll learn
about advanced multivariate statistical techniques. But you'll also learn about
non-statistical approaches such as interviews and discourse analysis.

The work you do in these modules will also connect up with other modules:

1. **Analysing Data** with, for example, **Psychobiology**

2. **Discovering Statistics** with, for example, **Developmental Psychology**

3. And **Quantitative and Qualitative Methods** with, for example, **Social
   Psychology**

Research methods doesn't happen in isolation, but it's connected with everything
else you do in your degree

### Why research methods?

The dominant approach to **training psychologists** is the **scientist
practitioner model**. _Doing research_ is seen as integral to this approach!

Just like _medical doctors_ not only deliver treatments but also _develop_
treatments, the same goes for _psychologists._ As a _psychologist_ you want to
do _what works_ and being able to **read**, **critique**, and **conduct**
research will help you know _what works_ and allow you to develop
_evidence-based care_

Even for those that don't become psychologists research methods is still useful
and the skills you'll learn in your research methods courses will prepare you
for careers in a range of industries including working as a data scientist, in
consultancy, and the civil service and government.

## Structure of this module

Like most research methods modules PAAS is made up of **three** main activities:

1. Weekly lectures

   - One hour each week.

   - Cover research methods, statistics, and theory

   - Note that you'll also have a special lecture one evening in Week 6 that
     will cover research ethics. Check your timetable!

2. Tutorials/Practical preparation homework

   - About an hour a week.

   - Done online or in `R` as preparation for the practical class

3. Practical classes

   - Two hours a week

   - Hands on experience with the `R` programming language

### Assessment Structure

For the assessment there's a 50/50 split between **coursework** and the
**exam**.

There a four parts to the course work. Pay special attention to this section,
because the list on **Sussex Direct** can be a little confusing.

1. **Computer Based Exam** worth 10% due in approximately Week 8

   This will cover the material from the _ethics lecture_ in Wk 6

2. **Computer Based Exam** worth 40% (see below about due date)

   This refers to the weekly `R` assessments that you'll do. You'll find out
   more about these in the practical classes. It is **very important** to note
   that this isn't a single assessment due at the end of term. Rather these are
   continual weekly assessments. Make sure you keep up with them, because doing
   them will really help you acheive a good grade.

3. **Report** worth 40% due in approximately Wk 9

4. **Portfolio** with 10% (listed on Sussex Direct as due in Wk 11)

   This refers to the 20 credits worth of research participation that you're
   required to do as part of the course. To find out more about this follow the
   big yellow link labelled **Research Participation/Sona**.

   The due date is the _final day_ you can complete research participation.
   **Don't wait until the last week** to do it, because there'll probably be no
   studies left to take part in. _These are easy marks so don't miss out on
   getting them._

## Lecture topics

The rough outline of the lecture schedule is as follows:

```{r}
#| echo: false


tibble::tribble(
~Week, ~Topic,
1, 'Introduction to PAAS',
2, 'What is this thing called "Science"',
3, 'Approaches to research',
4, 'Measurement and variables',
5, 'Open Science',
6, 'Samples, populations, and distributions',
7, 'Descriptive statistics and the sampling distribution',
8, 'Distributions, functions, and transformations',
9, 'Summarising data in tables and plots',
10, 'Probability theory',
11, 'Recap and exam guide') |>
  knitr::kable(align = c('l','l'))
```

The first set of lectures will cover _big picture_ ideas. These lectures will
probably be most useful in helping you to prepare for the report.

In these lectures we'll talk about issues like:

- What are scientific theories

- What issues do we need to consider when we're measuring things?

- What does it mean to operationalise our variables?

- What are different approaches you can take when conducting a study?

- What are some sources of bias in psychology studies and publishing of
  psychology studies and how might we be able to ameliorate some of these
  biases?

The second set of lectures are all about preparing you for learning about
statistics and working with data.

In these lectures you'll learn the underlying theory of _statistical testing_.
You'll learn how to _reason about statistics and data_, and the relationship
between scientific hypotheses and statistical hypotheses.

Doing statistics isn't like following a recipe. It's **not** about just picking
the **"correct"** statistical tests out of a list. It involving thinking about
what you want to know, why you want to know it, and how statistics can help you
to know it. So we spent a bit of time this term just learning about this
reasoning before you actually learn about statistical tests next term.
