In the previous lecture, we started talking about something called the
**standard error of the mean**. We said that the **standard error of the mean**
told us something about how spread out the **sample means** would be if we took
multiple samples from the same population. That is, the **standard error of the
mean** if the **standard deviation** of something called the **sampling
distribution**. But before we can talk about the **sampling distribution**, we
need to talk about **distributions** more generally, where they come from, and
why they look the way that they do. That means that for now, we'll put the
**standard error of the mean** aside, but we will return to it later.

Up until now, we've skirted around the idea of distributions. We've
looked at histograms of data, but we haven't really talked much about their
shape. It turns out that there are some shapes that we'll come across very
often, and some of these shapes have properties that will make them very useful
for statistics. But before we can get to that, we need to first understand where
these shapes come from---that is, why distributions have the shape they do---and
some language for describing these shapes. We'll start off with the simplest
distribution, the **binomial distribution**, before moving on to the **normal
distribution**.

## The binomial distribution

To understand what the **binomial distribution** is, and where it comes from
we'll do a little _thought experiment_. In our thought experiment, we'll take a
coin, and we'll flip it. When we flip a coin, one of two outcomes is possible.
Either the coin will land showing heads, or it will land showing tails. We can
say that there are two possible events or two possible sequences of events
(_sequences_ will make more sense when we add more coins) that can happen when
we flip a coin.

But now let's make it a little more complicated. Let's flip two coins. Now
there's a greater number of possible sequences. We can list them:

1. The first coin shows heads, and so does the second (HH),
2. The first coin shows heads and the second shows tails (HT),
3. The first coin shows tails and the second shows heads (TH),
4. and the first coins shows tails and the second shows tails (TT)


Now there are four possible sequences. Let's count up the **number of
sequences** that lead to 0 heads, one head, two heads, etc. If we do this, we'll
see that one sequence leads to 0 heads (TT). Two sequences lead to 1 head (HT,
and TH). And one sequence leads to 2 heads (HH).

Let's now add more coins. Things will get trickier from here, because the number
of sequences rapidly goes up. With three coins there will be `r 2^3` possible
sequence. With four coins there will be `r 2^4` sequences. And with five coins
there will be `r 2^5` possible sequences. To make things easier, we'll draw a
plot. First, we'll draw out a plot to trace the sequences. We'll use different
coloured dots to indicate heads and tails. We can do this in the form of a
branching tree diagram shown in @fig-sequences.

Once we've visualised the sequences, it's easy to count up how many sequences
result in 0 heads, one head, two heads etc. For this, we'll make a frequency
plot, or histogram, just like we've seen before. On the x-axis, we'll have the
number of heads. And on the y-axis, we'll have the number of sequences that
result in that number of heads. This frequency plot is shown in @fig-binomdist.


```{ojs}
//| label: fig-sequences
//| fig-cap: Possible sequences after ${coins - 1} coin flips
viewof sequences = {
  const div = document.createElement("div");
  div.value = new vega.View(parsedSpec).initialize(div).run();
  return div;
}
```

```{ojs}
//| panel: fill
//| label: fig-binomdist
//| fig-cap: Distribution of number of heads after ${coins - 1} coin flips
viewof coin_flip_dist = Plot.plot({
  y: {
    label: "Sequences",
    grid: true
  },
  x: {
    label: "Number of heads"
  },
  marks: [Plot.barY(n_heads, { x: "x", y: "y" }), Plot.ruleY([0])]
})
```

```{ojs}
//| width: 100%
//| panel: input
viewof coins = htl.html`<input style="width:300px" type="range" id="coins" min="1" max="7" value="1" class="form-range">`

coins_label = htl.html`<label for="coins" class= "form-label" width="100%">Number of coin flips: ${
  coins - 1
}</label>`
```


Neither @fig-sequences nor @fig-binomdist look very interesting at the moment.
But that's because we haven't started flipping any coins. Move the slider above
so that you can see what happens after we've flipped one coin, two coins, three
coins etc. 

Increasing the number of coin flips increases the number of possible
sequences. And it changes the number of ways of getting one head, two heads,
three heads and so on. Notice that as you adjust the slider and add
more and more coin flips, the frequency plot takes on a characteristic shape. You
can mathematically model the shape of this plot using a **binomial
distribution**.

The binomial distribution is just an **idealised representation** of the
**process** that **generates** sequences of heads and tails when we flip a coin.
It tells us that if we flip a coin a certain number of times, and that coin
lands head 50% of the time, then there is a particular number of sequences that
will produce 0 heads, 1 head, 2 heads etc. And the shape of the distribution
tells us how many sequences will produce each of those outcomes. We'll encounter
the binomial distribution again when we talk about probability, but for now
hopefully the intuition make sense.

I said it's an **idealised** representation, but we can also see that
stereotypical shape being produced by natural processes. One natural process
that gives rise to this shape is the "bean machine". In a bean machine, small
steel balls fall from the top of the device to the bottom of the device. On
their way down, they bump into pegs. When one of the balls hits a peg, it has a
roughly equal chance of bouncing off to the left or the right. At the bottom of
the device are equally-spaced bins for collecting the balls. If enough balls are
dropped into the device, then the **distribution** of balls across the bins will
start to take on the shape of the **binomial distribution**. Very few balls will
be at the far edges, because this would require the balls to bounce left or
right every time. Instead, most of the balls will tend to be clumped somewhere
near the middle. You can see an example of a "bean machine" in @fig-bean.

```{r}
#| echo: false
#| message: false
#| fig.cap: Example of the bean machine
#| label: fig-bean
require(htmltools)
vid <- tags$video(width = "100%", controls = NA)
# source <- tags$source(src = here::here("images","bean_machine.webm"), type = "video/webm")
data <- xfun::embed_file(here::here("images/bean_machine.webm"))[2]$attribs$href
source <- tags$source(src = data, type = "video/webm")
vid$children = source
vid
```

Flipping coins might seem a long way off from anything you might want to study in
psychology. However, the **shape** of the binomial distribution, might be
something you're more familiar with. This characteristic **bell shape** is also
something we see in the **normal distribution**. And it's the **normal
distribution** which we'll turn our attention to next.

## The normal distribution

The **normal distribution** has a similar shape to the **binomial
distribution**; however, there are a few key differences. First, the binomial
distribution is **bounded**. One end represents 0 heads. And the other end
represents all heads. That is, the distribution can only range from 0 to n
(where n is the number of coins). The **normal distribution**, however, is
unbounded. It can range from **positive infinity** to **negative infinity**. The
second difference is that for the binomial distribution, the steps along the
x-axis are **discrete**. You can have 0 heads, 1 head, 2 heads etc. But you
can't get 1.5 heads. But for the **normal distribution**, the steps are
**continuous**.

The **normal distribution** is a mathematical abstraction, but we can use it as
a **model** of real-life frequency distributions. That is, we can use it as a
model of **populations** that are produced by certain kinds of natural
processes. Because normal distributions are unbounded and continuous, nothing,
in reality, is normally distributed. For example, it's impossible to have
infinity or negative infinity of anything. This is what is meant by an
**abstraction**. But natural processes can give rise to frequency distributions
that look a lot like normal distributions, which means that normal
distributions can be used as a model of these processes. Before 
we talk more about the shape of the normal distribution, let us first 
examine some processes that give rise to normal distributions.

### Processes that produce normal distributions 

To see how a natural process can give rise to a normal distribution, let us
play a board game! We won't play it for real, but we'll *simulate* it. 

In this game, each player rolls the dice a certain number of times, and they
move the number of spaces indicated by the dice. Not that dissimilar to what
you'd do in a game of monopoly, or a similar board game! For example, if a
player rolled the die three times, and they got 1, 4, 3, then they would move
`r 1 + 4 + 3` (1 + 4 + 3 = `r 1 + 4 + 3`) spaces along the board. At the end of
one round of rolls we can take a look at how far from the start each player is.
And we can draw a histogram of this data. You can explore the simulation in
@exm-dice-plus.

::::{.callout-tip icon="false" appearance="simple"}

:::{#exm-dice-plus}
#### Dice game simulation (adding)
:::

In the simulation, you can set the number of players and the number of rounds
they play. We'll start off by having 10 players, and each player will roll 1
dice/play 1 round. Next try setting the _Number of rounds_ to **2** and then
increasing the _Number of players_. What do you notice? You can also try
increasing the _Number of rounds_ so that it's more than **2**. What do you
notice?



```{ojs}
//| echo: false
//| label: fig-dice-dist1
//| fig-cap: Distribution of players' position from the starting point
simple_dice_plot = Plot.plot({
  x: {
    label: "places from start",
    domain: d3.sort(dicedata, (d) => d.value).map((d) => d.value)
  },
  y: {
    label: "number of players"
  },
  marks: [
    Plot.barY(dicedata, {
      x: "value",
      y: "count",
      sort: { x: "y", reverse: true }
    }),
    Plot.ruleY([0])
  ]
})
```

```{ojs}
//| echo: false
//| panel: input
//| layout-ncol: 2
viewof n_dice = Inputs.range([1, 2000], {
  step: 1,
  value: 1,
  label: "Number of rounds"
})
viewof n_players = Inputs.range([1, 10000], {
  step: 1,
  value: 10,
  label: "Number of players"
})
```

::::


When you have enough players, and they play enough rounds, then the
distribution shown in @fig-dice-dist1 starts to take on the shape of a **normal
distribution**. What is it about the dice game that gives rise to the normal
distribution shape? In the dice game each players position is determined by
**adding up** the value of each roll. This **adding up** is the key feature
that determines the shape of the distribution of the players' scores. Notice
that this is also what we did with coin flips. We **added up** the number of
heads. It doesn't matter whether we're adding up the number of heads or the
outcomes of dice rolls. What matters is that each final value (number of
heads or places from start) is determined by adding up numbers.

We can imagine other processes that might work similarly to the dice game. For
example, a **developmental process** might work similarly. Let's say that we
have a developmental process like height. At each **point in time** some
**value** (growth) can be **added** on to a person's current height. That is,
at each time point a person can **grow** by some amount just like each players'
score can increase by some amount with each dice roll. If you have enough
people (players), and some time for growth to happen (rolls), then the
**distribution** of **height** will start to look like a normal distribution.


#### Processes that don't produce normal distributions

I've said that the key factor that gives rise to the normal distribution is
that outcomes are determined by **adding** together numbers^[Or
**subtracting**, because **subtracting** is just adding negative numbers]. We
can test out this idea by changing the rules of the dice game so that instead
of **adding** together numbers we'll **multiply** them.

In @exm-dice-mult we've changed the rules so that each players' score is
determined by multiplying together the values of their rolls. For example,
under the new rules, if a player rolled 1, 4, 3 then their score would be 12 (1
× 4 × 3 = `r 1 * 4 * 3`). Explore what happens to the shape of the distribution
under these new rules.



:::{.callout-tip icon="false" appearance="simple"}

:::{#exm-dice-mult}
#### Dice game simulation (multiplying)
:::

As with the previous simulation you can set the number of players and the
number of rounds they play. In the previous simulation when you had many
players that played a lot of rounds you saw a distribution that was shaped like
a **normal distribution**. For this simulation first try setting the number of
players to the maximum (400), and then slowly increase the number of rounds
they play from 1, to 2, to 3, etc. Notice how the shape of the distribution
differs from the shape of the normal distribution.

```{ojs}
//| echo: false
dice_plot_mult = Plot.plot({
  x: { label: "places from start" },
  y: { label: "number of players" },
  marks: [
    Plot.rectY(d, Plot.binX({ y: "count" }, { x: "x" })),
    Plot.ruleY([0]),
  ],
});
```



```{ojs}
//| echo: false
//| panel: input
//| layout-ncol: 2
viewof n_dice_mult = Inputs.range([1, 20], {
  step: 1,
  value: 6,
  label: "Number of rounds"
})
viewof n_players_mult = Inputs.range([5, 400], {
  step: 1,
  value: 200,
  label: "Number of players"
})
```



::::


```{r}
#| echo: false
GBP <- scales::dollar_format(prefix = "£")
```

In the new version of the dice game, the distribution of scores doesn't look
like a normal distribution. After only a few rounds you start seeing a few
players with very high scores, but most of the players are clumped around near
the start. You might notice that this plot looks a little bit like the plot we
saw in Lecture 6 of the Average Annual Salary across a set of 79 countries.
This is maybe not a total coincidence. Can you think of a process that might
operate with similar rules as this game? How about interest, or returns on
investments?

When you get a 10% annual return on an investment, and you start with £10, then 
after 1 year you'll have 10 × 1.10 or `r GBP(10 * 1.10)`. After two years, you'll 
have (10 × 1.10) × 1.10 or `r GBP((10 * 1.10) * 1.10)`. 

Given this, it's not surprising to see wealth distributions (whether across
countries in the form of national salaries or within countries) to be heavily
**skewed** in a manner resembling the outcomes of the dice game in
@exm-dice-mult. It's simply a consequence of the rules of the game where wealth
accumulates more wealth by multiplication! And in our dice game, this
**skewed** distribution happens even though each player is rolling the same
set of dice. A lucky few rolls at the start of the game and then some players
get so far ahead that they'll never get caught by the remaining players stuck
down near the bottom.

In psychology, we won't study many processes that grow like this, 
but it is an interesting exercise thinking about these processes. 


### Describing normal distributions

The normal distribution has a characteristic bell shape, but not all normal
distributions are identical. They can vary in terms of where they're centered
and how spread out they are. We've already talked about measures of central
tendency and measures of spread. In particular, we've talked about the mean and
standard deviation, and these are what we use to describe the shape of the
normal distribution.

The mean ($\mu$) determines where the centre is, and the standard deviation
($\sigma$) determines how spread out it is. Use @exm-norm to explore the
**normal distribution**.

::::{.callout-tip icon="false" appearance="simple"}

:::{#exm-norm}
#### Explore the normal distribution
:::

Use the sliders to make adjustments to $\mu$ and $\sigma$ and see how the
plot of the normal distribution changes.


```{ojs}
normal_plot_output = Plot.plot({
  x: {
    grid: true,
    domain: [-4, 4]
  },
  y: {
    grid: true,
    domain: [0, 0.8]
  },
  marks: [
  // Plot.line(
  //   normal_plot(
  //     mean_value - sd_value * s > -4 ? mean_value - sd_value * s : -4,
  //     mean_value + sd_value * s < 4 ? mean_value + sd_value * s : 4,
  //     mean_value,
  //     sd_value
  //   ),
  //   {
  //     x: "x",
  //     y: "y",
  //     strokeWidth: 1,
  //     fill: "blue",
  //     opacity: show ? 0.5 : 0
  //   }
  // ),
  // Plot.line(fill_limits(s), {
  //   x: "x",
  //   y: "y",
  //   fill: "blue",
  //   strokeWidth: 1,
  //   opacity: show ? 0.5 : 0
  // }),
    Plot.line(normal_plot(-4, 4, mean_value, sd_value), {
      x: "x",
      y: "y",
      strokeWidth: 4
    }),
    Plot.ruleY([0])
  ]
})
```

```{ojs}
normal_sliders = htl.html`${mean_value_slider}${sd_value_slider}`
```

```{ojs}
texmd`This plot shows a normal distribution with a mean ($\mu$) of ${mean_value} and a standard deviation ($\sigma$) of ${sd_value}.`
```

::::

We'll come to this point later again, but I'll mention it briefly here. One
important thing to note is that when we make adjustments to the $\mu$ and
$\sigma$ of the **normal distribution**, the **relative position of points (in
units of $\sigma$**) doesn't change, even though the absolute position of
points on the plot do change. 

<!--

This is rather a mouthful. So let's break down what it actually means.  If we
take a normal distribution with $\mu$ = 0 and $\sigma$ = 1, then about 68% of
the plot lies between -1 and 1. In units of $\sigma$ this means that about 68%
of the plot lies between -1 $\sigma$ and +1 $\sigma$ of $\mu$. If we were to
adjust $\mu$ and $\sigma$ to, for example, $\mu$ = 2, and $\sigma$ = 4, then
68% of the plot would now lie between `r 2 - 4` and `r 2 + 4`. But in units of
$\sigma$, -2 is just -1 $\sigma$ from $\mu$ (2 [$\mu$] - 4 [$\sigma$] = 2), and
6 is just +1 $\sigma$ from $\mu$ (2 [$\mu$] + 4 [$\sigma$] = 6). In fact, no
matter how we change $\mu$ and $\sigma$, 68% of the plot will always be between
$\mu$ - $\sigma$ and $\mu$ + $\sigma$. This will be the case with all normal
distributions. 

-->

But sometimes we see distributions that aren't normal distributions. They've
departed from this **normal** distribution shape in some way. There are
specific terms for describing departures from the shape we see in the normal
distribution. And it's these terms that we turn our attention to next.


### Describing departures from the normal distribution.

When we looked at the example of the game where a players score was based on
**multiplying** together the dice rolls, it produced a distribution that was
**skewed**.

**Skew** is actually a technical term that describes one way in which a
distribution can _deviate_ from the shape we see with _normal distributions_.
The _normal distribution_ is **symmetrical**, but a **skew** distribution is
not. A left-skewed distribution has a longer _left tail_, and a
**right-skewed** distribution has a longer _right tail_. Use @exm-skew to
explore **skewness**.

::::{.callout-tip icon="false" appearance="simple"}

:::{#exm-skew}
#### Explore Skewness
:::

Use the slider to make adjustments to the skewness of the distribution. Slide it
to the *left* to add **left-skew** and slide it to the *right* to add
**right-skew**. See how the distribution changes. When the distribution has no
skew then it is symmetrical just like a normal distribution. 

```{ojs}
skew_normal_plot_output = Plot.plot({
  x: {
    grid: true
  },
  y: {
    grid: true,
    domain: [0, 0.8]
  },
  marks: [
    Plot.line(skew_normal_plot(-4, 4, skew), {
      x: "x",
      y: "y",
      strokeWidth: 4
    }),
    Plot.ruleY([0])
  ]
})
```

\
```{ojs}
viewof skew = Inputs.range([-10, 10], { label: "Skewness", step: 0.5 })
```

```{ojs}
skew_desc = md`This distribution has a **skewness** of ${skew}. It is ${
  skew == 0 ? "symmetrical" : skew < 0 ? "left-skewed" : "right-skewed"
}.`
```

::::


Apart from **skew**, deviations from the **normal** distribution can occur when
a distribution either has fatter or skinnier **tails** than the normal
distribution. The _tailedness_ of a distribution is given by its **kurtosis**.
The kurtosis of a distribution is often specified with reference to the **normal
distribution**. In this case, what is being reported is **excess** kurtosis. A
distribution with **positive excess kurtosis** has a higher kurtosis value than
the normal distribution, and a distribution with **negative excess kurtosis**
has a lower kurtosis value than the normal distribution.

Distributions with no excess kurtosis are called **mesokurtic**. Distributions
with negative excess kurtosis are called **platykurtic**. And distributions
with positive excess kurtosis are called **leptokurtic**.

In your research methods courses, you probably won't come across many
distributions that have negative excess kurtosis. However, the distribution
that describes the outcomes of a single dice roll is one such distribution. You
will encounter distributions with positive excess kurtosis more often. In
particular, the _t_-distribution, a distribution with positive excess
kurtosis, will be used in several of the statistical procedures that you will
learn about next year. You can use @exm-kurt to explore excess kurtosis.

::::{.callout-tip icon="false" appearance="simple"}

:::{#exm-kurt}
#### Explore kurtosis
:::

Use the slider to adjust the excess kurtosis. A standard normal distribution
with an excess kurtosis of 0 is shown for reference.


```{ojs}
kurtosis_plot = Plot.plot({
  x: {
    grid: true,
    domain: [-3, 3]
  },
  y: {
    grid: true,
    domain: [0, 0.4]
  },
  marks: [
    true
      ? Plot.line(dists.standard_normal, {
          x: "x",
          y: "y",
          stroke: "grey",
          strokeWidth: 1
        })
      : null,
    Plot.line(dists[Object.keys(kurtosis)[kurtosis_value]], {
      x: "x",
      y: "y",
      strokeWidth: 4
    }),
    Plot.ruleY([0])
  ]
})
```

```{ojs}
//|echo: false
//| panel: input
viewof kurtosis_value = Inputs.range([0, 6], {
  step: 1,
  value: 2,
  label: "Excess kurtosis",
  format: (d) => kurtosis_values[d]
})
```

```{ojs}
kurtosis_desc = md`This distribution has an excess kurtosis of ${
  kurtosis_values[kurtosis_value]
}. ${
  kurtosis_values[kurtosis_value] === 0
    ? "It is a **Mesokurtic distribution**."
    : kurtosis_values[kurtosis_value] > 0
    ? "It is a **Leptokurtic distribution**."
    : "It is a **Platykurtic distribution**."
}`
```

::::


## Distributions and samples 

Now that we've talked a little bit about samples and populations (in Lecture 6
and Lecture 7) and we've also covered distributions, we're going to start
putting these ideas together. We've seen that whenever we look at the
distribution of values where the values are produced by **adding up numbers**
we got something that looked like a normal distribution.

When we covered the **sample mean** in Lecture 6, the formula was as shown
in @eq-mean, below:

$$\bar{x}=\frac{\displaystyle\sum^{N}_{i=1}{x_i}}{N}$$ {#eq-mean}

This formula can be re-written as shown in @eq-mean-sum, below:

$$\bar{x}={\displaystyle\sum^{N}_{i=1}{\frac{x_i}{N}}}$$ {#eq-mean-sum}

What @eq-mean-sum makes clear is that calculating a mean is just **adding up
numbers**. Now let's think about taking lots of samples from a population. And
for each sample, we calculate the sample mean. If we had to plot these sample
means, then what would the distribution look like? We can try it out. Let's
say that I have a population with a mean of 100 and a standard deviation of
15. From this population I can draw samples. Let's say that each sample will be
`r 5^2` values (that is, the sample size will be `r 5^2`). After I've
collected my sample I'll work out the sample mean, and I'll do this 100,000
times and plot the results. You can see this plot in @fig-sampling-r.

```{r}
#| label: fig-sampling-r 
#| fig.cap: Sample means from 100,000 samples (sample size = 25)
#| message: false
require(tidyverse)

mean <- 100
sd <- 15
n <- 5^2

samp <- \(x) rnorm(n = n, sd = sd, mean = mean) |> mean() 

d <- parallel::mclapply(1:100000, samp) |> unlist() 
df <- tibble::tibble(x = d)

ggplot(df, aes(x = x)) + 
geom_histogram(aes(y = ..density..), bins = 100) +
geom_function(fun = \(x) dnorm(x, mean,  sd / sqrt(n)), color = "red") +
theme_minimal() + 
labs(x = "sample mean") +
scale_y_continuous(breaks = NULL, name = "count") +
theme(panel.grid.major.x = element_blank()) +
theme(panel.grid.minor.x = element_blank())
```


As you can see in @fig-sampling-r the distribution of sample means is shaped
just like a normal distribution. This is exactly what you'd expect. To make it
clear that it's shaped like a normal distribution, in <font color="red">red</font>
I've drawn the outline of a normal distribution. This normal distribution has
a mean of 100, which is just what the mean of our population is. It also
has a standard deviation of `r 15 / sqrt(25)`. This distribution is 
an example of **the sampling distribution of the mean**. This means that the
standard deviation of this distribution is the **standard error of the mean**
because the **standard error of the mean** _is_ just the **standard deviation
of the sampling distribution of the mean**.

### The central limit theorem

Before we get to how we work out the **standard error of the mean** I just want
to assure you of something. You might think that in the example above the
sampling distribution of the mean has a normal distribution **because we're
drawing samples from a population that has a normal distribution**. That is,
you might think that because the parent distribution has a normal distribution,
the sampling distribution will have a normal distribution. 

But this is not the case. As long as we're dealing with the _sampling
distribution of the mean_ (that is, a value we calculate by adding up numbers)
then we'll see a sampling distribution that has the shape of a normal
distribution. You can explore this fact in @exm-central-limit


::::{.callout-tip icon="false" appearance="simple"}

:::{#exm-central-limit}
#### The Central Limit Theorem
:::

You can use the selector to select the population distribution. A histogram
that matches the shape of the population distribution is shown in
@fig-pop-central-limit.

```{ojs}
import {viewof poptype} from "@ljcolling/central-limit"
import {viewof n} from "@ljcolling/central-limit"
import {pop_hist_plot} from "@ljcolling/central-limit"
import {sampling_distrubution} from "@ljcolling/central-limit"
```

```{ojs}
viewof poptype

```

```{ojs}
//| fig-cap: Distribution of the population
//| label: fig-pop-central-limit
pop_hist_plot
```


After you've selected the population distribution, then you can draw samples
from that distribution. You can set the **sample size** for each sample you
draw. We'll always draw 50,000 samples, because that should be enough for the
shape of the sampling distribution to become clear. You can see a plot of the
sampling distribution of the mean in @fig-sampling-central-limit. When the
sample size is **1** then the sampling distribution should look just like the
population distribution. But try increasing the sample size and see what
happens. What do you notice?

```{ojs}
//| fig-cap: Sampling distribution of the mean (50, 000 samples)
//| label: fig-sampling-central-limit
sampling_distrubution
```

```{ojs}
//| panel: input
viewof n
```

When the sample size increases then the sampling distribution takes on the
shape of a normal distribution. The normal distribution that the sampling
distribution approximates is also shown in @fig-sampling-central-limit as
a solid <font color="blue">blue</font> line^[Note that the sampling distribution
and the normal distribution shown in the plot won't always match perfectly 
even when the sample size is large. This is because the normal distribution
is an idealisation of what happens if you collect an _infinite_ number of samples.
But we're not collecting an _infinite_ number. We're only collecting 50,000].

:::

<!-- TODO: FINISH OFF AND THEN DO FORMULA FOR STANDARD ERROR -->

<!--
As long as the sample size is sufficiently large then the **sampling
distribution (of the mean)** will have a normal distribution. 

Note also that the **sampling distribution of the mean** isn't the same as the
distribution of the **population**. The population distribution can have any
shape, as long as the samples are sufficiently large then the **sampling
distribution of the mean** will have a normal distribution.


The words **sufficiently large** appear to be doing a lot of work in what I've
just said. So what exactly does **sufficiently large** mean? Like a lot of
things in statistics, it depends! Specifically, it depends on the population
distribution. In the population is perfectly normally distributed then a sample
size of 1 is sufficient. But if it's not, then you'll need bigger samples.
You'll often hear people suggest heuristics or rules of thumb for how big the
samples need to be. But I won't. I won't because I think people often end up
mistaking rules of thumb or actually hard and fast rules when they're not. So
you'll have to make do with the ambiguity of **sufficiently large**!

-->

## The standard error of the mean

<!--
We've now had the chance to see a few different sampling distributions. And
we've seen that sampling distributions of the mean look like normal
distributions that have a mean that is equal to the population mean and that
they also have a standard deviation (the **standard error of the mean**). But
how do we work out the **standard error of the mean**?

In the Lecture 7, I asked you to try think of two situations where all the samples drawn
from a population will have idenitical means---or put another way, two situations where
the standard error of the mean would be 0. 
-->

In Lecture 7 we started talking about the spread of **sample means** around the
**population mean** that occurs when we repeatedly draw samples from the same
population. I showed you [an
example](https://paas.netlify.app/lectures/week07/handout/#fig-se) where the
**average deviation** was small (that is, _on average_ the individual sample
means were close to the population mean). And I showed you an example where the
**average deviation** was larger (that is, _on average_ the individual sample
means were further away from the population mean). We also learned in Lecture 7
that **variance** is a measure of **average (squared) deviation**.

I ended the lecture by asking you to think of two situations where the 
average deviation (of sample means around the population mean) would be
small or 0. If you managed to think of those two situations, then great!
But if you didn't, then here they are:

First, a feature of populations that will make the average deviation
of sample means around the population mean be equal to 0. If the 
variance in the population is 0 then all sample means will be identical 
to the population mean. Why? Because if the variance in the population
is 0 then all members of the population will be identical. The 
typical value (the mean) any sample will have to be identical to the 
population mean, because there's only one value anything can take. 

Second, a feature of the samples that will make the average deviation of
samples means around the population mean be equal to 0. If the sample size is
so large that it includes the entire population then the sample mean would, by
definition, be equal to the population mean. So if all your samples are this
large then all the sample means will be equal to the population mean. And this
means that the average deviation of sample means around the population would be
equal to 0.

This means that if we want to compute the **average (squared) deviation** of
sample means around the population mean then it will depend on two factors.
First, when the **variance of the population** ($\sigma^2$) is small then the
_variance of sample means around the population mean_ will be small.
Conversely, when the **variance of the population** is larger then we'd expect
the _variance of sample means around the population mean_ to be larger. Second,
if the sample size (n) is large then the _variance of the sample means around the
population mean_ will be small. Conversely, when the sample is small then
we'd expect the _variance of sample means around the population mean_ to be
larger. 

The only way to combine $\sigma^2$ and $n$ that matches these observations
is as shown in @eq-sem-squ, below:


$$\frac{\sigma^2}{n}$${#eq-sem-squ}


This formula gives us the _variance of the sample means around the population
mean_. But I said that the **standard error of the mean** is the **standard
deviation** of the sample means around the population mean. That means that we
just need to take the square root of the formula in @eq-sem-squ. And because we
don't actually know the value of $\sigma$ (the variance of the population) and
only know our **estimate of it**, the sample variance (s), we need to replace
$\sigma$ with s. This gives us the equation in @eq-sem, below: 


$$\frac{s}{\sqrt{n}}$${#eq-sem}

This is exactly the formula for **the standard error of the mean**.

Now this is, admittedly, a fairly long winded way to get to what is essentially
a very simple formula. However, as I have alluded to several times, the
**standard error of the mean** is a fairly misunderstood concept. Part of the 
reason for this is that it's often just presented as a formula. I hope that
getting there the long way has helped you to build a better intuition of
what the standard error of the mean actually is.





<!-- Dependencies {{{ -->

```{ojs}
n_heads = jstat(0, coins - 1, coins + 1 - 1)[0].map((v) => {
  return {
    x: v,
    y: jstat.binomial.pdf(v, coins - 1, 0.5) * 2 ** (coins - 1)
  };
})
```

```{ojs}
vega = require("https://cdn.jsdelivr.net/npm/vega@5/build/vega.js")
```

```{ojs}
coin_data = [
  { name: "START", id: 1, parent: "", color: "red" },
  ...d3.range(2, 2 ** coins).map((i) => {
    return {
      name: i % 2 ? "T" : "H",
      id: i,
      parent: Math.floor(i / 2)
    };
  })
].map((x) => {
  let colors = { H: "black", T: "white", START: "red" };
  x.color = colors[x.name];
  return x;
})
```

```{ojs}
parsedSpec = {
  return vega.parse(spec3);
}
```

```{ojs}
spec3 = {
  return {
    $schema: "https://vega.github.io/schema/vega/v5.0.json",
    padding: 0,
    width: 500,
    height: 100,
    layout: {
      padding: 0,
      columns: 1
    },
    marks: [
      {
        type: "group",
        encode: {
          update: {
            width: {
              value: 1000
            },
            height: {
              value: 130
            }
          }
        },
        data: [
          {
            name: "tree",
            values: coin_data,
            transform: [
              {
                type: "stratify",
                key: "id",
                parentKey: "parent"
              },
              {
                type: "tree",
                method: "tidy",
                size: [500, 200],
                as: ["x", "y", "depth", "children"]
              }
            ]
          },
          {
            name: "links",
            source: "tree",
            transform: [
              {
                type: "treelinks",
                key: "id"
              },
              {
                type: "linkpath",
                orient: "horizontal",
                shape: "line"
              }
            ]
          }
        ],
        scales: [
          {
            name: "color",
            domain: [0, 1, 2, 3, 4, 5],
            type: "sequential",
            range: "ramp"
          }
        ],
        marks: [
          {
            type: "path",
            from: {
              data: "links"
            },
            encode: {
              update: {
                path: {
                  field: "path"
                },
                stroke: {
                  value: "black"
                }
              }
            }
          },
          {
            type: "symbol",
            from: {
              data: "tree"
            },
            encode: {
              enter: {
                size: {
                  value: 50
                },
                stroke: {
                  value: "black"
                }
              },
              update: {
                x: {
                  field: "x"
                },
                y: {
                  field: "y"
                },
                fill: {
                  field: "color"
                }
              }
            }
          },
          {
            type: "text",
            from: {
              data: "tree"
            },
            encode: {
              enter: {
                text: {
                  field: "name"
                },
                fontSize: {
                  value: 0
                },
                baseline: {
                  value: "bottom"
                }
              },
              update: {
                x: {
                  field: "x"
                },
                y: {
                  field: "y"
                }
              }
            }
          }
        ]
      }
    ]
  };
}
```

```{ojs}
jstat = require("https://bundle.run/jstat@1.9.4")
```

```{ojs}
dicedata = {
  return d3.sort(
    dist.six_dice_roll_histogram(n_dice, n_players).counts,
    (d) => d.value
  );
}

```

```{ojs}
d = {
  return Array(n_players_mult)
    .fill(0)
    .map((x) => {
      return {
        x: Number(
          Array.from(dist.six_dice_roll(1, n_dice_mult)).reduce(
            (state, item) => state * item
          )
        )
      };
    });
}
```

```{ojs}
import { dist } from "@ljcolling/wasm-distributions"
```

```{ojs}
sd_value_slider = Inputs.range([0.5, 2], {
  value: 1,
  step: 0.25,
  label: htl.html`standard deviation <br />&#x3C3;`
})
```

```{ojs}
import { texmd } from "@kelleyvanevert/katex-within-markdown"
```

```{ojs}
mean_value_slider = Inputs.range([-3, 3], {
  value: 0,
  step: 0.25,
  label: htl.html`mean<br />  &#x3BC`
})
```

```{ojs}
sd_value = Generators.input(sd_value_slider)
```

```{ojs}
mean_value = Generators.input(mean_value_slider)
```

```{ojs}
// jStat.normal.pdf( x, mean, std )
normal_plot = (min, max, mean, sd) => {
  // jStat.normal.pdf(x, mean, sd)

  return d3.ticks(min, max, 501).map((v) => {
    return {
      x: v,
      y: dnorm(v, mean, sd)
    };
  });
}
```

```{ojs}
skew_normal_plot = (min, max, alpha) => {
  // jStat.normal.pdf(x, mean, sd)

  return d3.ticks(min, max, 201).map((v) => {
    return {
      x: v,
      y: dsn(v, alpha)
    };
  });
}
```

```{ojs}
dsn = (x, alpha) => {
  // set the defaults

  const xi = 0;
  const omega = 1;
  const tau = 0;

  let z = (x - xi) / omega;

  let logN = -Math.log(Math.sqrt(2 * Math.PI)) - 0 - Math.pow(z, 2) / 2;

  let logS = Math.log(
    jStat.normal.cdf(tau * Math.sqrt(1 + Math.pow(alpha, 2)) + alpha * z, 0, 1)
  );

  let logPDF = logN + logS - Math.log(jStat.normal.cdf(tau, 0, 1));

  return Math.exp(logPDF);
}
```

```{ojs}
// import jStat library
jStat = require("jStat")
```

```{ojs}
kurtosis = {
  return {
    uniform: -(6 / 5),
    raised_cosine: (6 * (90 - Math.PI ** 4)) / (5 * (Math.PI ** 2 - 6) ** 2),
    standard_normal: 0,
    t_dist30: 6 / (30 - 4),
    t_dist20: 6 / (20 - 4),
    t_dist10: 6 / (10 - 4),
    t_dist7: 6 / (7 - 5),
    t_dist5: 6 / (5 - 4)
  };
}
```

```{ojs}
kurtosis_values = Object.values(kurtosis).map((v) => Math.round(v * 100) / 100)
```

```{ojs}
dists = {
  return {
    raised_cosine: d3.ticks(-3, 3, 500).map((v) => {
      return {
        x: v,
        y: dist.raised_cosine(v, 0, 2.5)
      };
    }),
    standard_normal: d3.ticks(-3, 3, 500).map((v) => {
      return {
        x: v,
        y: dnorm(v, 0, 1)
      };
    }),
    t_dist30: d3.ticks(-3, 3, 500).map((v) => {
      return {
        x: v,
        y: dist.dt(v, 30)
      };
    }),
    t_dist20: d3.ticks(-3, 3, 500).map((v) => {
      return {
        x: v,
        y: dist.dt(v, 20)
      };
    }),
    t_dist10: d3.ticks(-3, 3, 500).map((v) => {
      return {
        x: v,
        y: dist.dt(v, 10)
      };
    }),
    t_dist7: d3.ticks(-3, 3, 500).map((v) => {
      return {
        x: v,
        y: dist.dt(v, 7)
      };
    }),
    t_dist5: d3.ticks(-3, 3, 500).map((v) => {
      return {
        x: v,
        y: dist.dt(v, 5)
      };
    }),

    uniform: d3.ticks(-2.1, 2.1, 500).map((v) => {
      return {
        x: v,
        y: dist.dunif(v, -2, 2)
      };
    })
  };
}
```

<!--
```ojs
viewof show = true//Inputs.toggle({ label: "Show coverage", value: false })
```
-->

<!--
```ojs
viewof s = 1//= Inputs.range([0, 3], { step: 0.5 })
```
-->

<!--
```ojs
fill_limits = (mult) => {
  let s = sd_value * mult;
  return [
    { x: mean_value - s > -4 ? mean_value - s : -4, y: 0 },
    normal_plot(
      mean_value - s > -4 ? mean_value - s : -4,
      mean_value - s > -4 ? mean_value - s : -4,
      mean_value,
      sd_value
    )[0],
    normal_plot(
      mean_value + s < 4 ? mean_value + s : 4,
      mean_value + s < 4 ? mean_value + s : 4,
      mean_value,
      sd_value
    )[0],
    { x: mean_value + s < 4 ? mean_value + s : 4, y: 0 }
  ];
}
```
-->


<!-- }}} -->


## Check your understanding

Use this quiz to make sure that you've understood the key concepts.

```{r}
#| echo: false
#| message: false

require(niceQuiz)

sem <- function(x) {
  s <- sd(x)
  n <- length(x)
  s / sqrt(n)
}

quiz(
  caption = "",
  question(
    "The mean of a set of numbers is 10, the variance is 25, and the sample size is 16. <br />What is the standard error of the mean?",
    numericanswer(sqrt(25)/sqrt(16)),
    type = "fillin")
    )
```

```{r}
quiz(
caption = "",
question(
  "A distribution with fatter tails than a normal distribution is called?",
  answer("Mesokurtic"),
  answer("Platykurtic"),
  answer("Leptokurtic", correct = TRUE)),

question(
  "A distribution with skewness of -1.5 is:",
  answer("Right-skewed"),
  answer("Left-skewed", correct = TRUE)),

question("The sampling distribution of the mean with have what kind of distributions?",
answer("A normal distribution", correct = TRUE),
answer("The same distribution as the population"),
answer("A distribution that is more skewed than the population distribution")
)





)

```




```{ojs}
function dnorm(x, mean, sd) {
  return jstat.normal.pdf(x, mean, sd);
}
```
